---
title: "Banded Dotterel Stable Isotopes"
subtitle: | 
  Exploration of dataset
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    code-fold: true
    code-tools: true
    self-contained: true
    highlight-style: github
    theme: Cosmo
execute:
  warning: false
  cache: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(cache = TRUE)
```

```{r, message=FALSE, results='hide', warning=FALSE, results='hide', include=FALSE, cache=FALSE}
## Prerequisites

### R packages
# Set a CRAN mirror before installing packages
options(repos = c(CRAN = "https://cloud.r-project.org"))
# -   The following packages are needed for analysis and can be easily installed from [CRAN](http://cran.r-project.org/) or GitHub by running the following code chunk:

# a vector of all the packages needed in the project
packages_required_in_project <- c("V8", "concaveman",
                                  "tidyverse",
                                  "readxl",
                                  "RMark",
                                  "RColorBrewer",
                                  "patchwork",
                                  "mapview",
                                  "lubridate",
                                  # "extrafont",
                                  "here",
                                  "DT",
                                  "leaflet",
                                  "sf",
                                  "leafpop",
                                  "tsibble",
                                  "corrplot",
                                  "gghalves",
                                  "gam",
                                  "pscl",
                                  "gamlss",
                                  "gt",
                                  "lme4",
                                  "ggpattern",
                                  "gtsummary",
                                  "effects",
                                  "lattice",
                                  "rptR",
                                  "partR2",
                                  "broom.mixed",
                                  "forcats",
                                  "glmmTMB",
                                  "ggmap",
                                  "MuMIn",
                                  "googlesheets4", 
                                  "sf",
                                  "sp", 
                                  "rnaturalearth", 
                                  "rnaturalearthdata", 
                                  "fuzzyjoin", "gstat", "conflicted")

# of the required packages, check if some need to be installed
new.packages <- 
  packages_required_in_project[!(packages_required_in_project %in% 
                                   installed.packages()[,"Package"])]

# install all packages that are not locally available
if(length(new.packages)) install.packages(new.packages)

# load all the packages into the current R session
lapply(packages_required_in_project, require, character.only = TRUE)

# set the home directory to where the project is locally based (i.e., to find 
# the relevant datasets to import, etc.
here::set_here()

# Use gs4_auth() to log in and authenticate with your Google account. 
# You can specify your email address to ensure the correct account is used. 
# Once authenticated, you won't need to select the account interactively when using read_sheet()
gs4_auth(email = "luke.eberhart@gmail.com")

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lmer", "lme4")
conflicted::conflict_prefer("accumulate", "purrr")
conflicted::conflict_prefer("ar", "brms")
conflicted::conflict_prefer("Beta", "brms")
# conflicted::conflict_prefer("calibration", "rstan")
conflicted::conflict_prefer("area", "MASS")
```

```{r, message=FALSE, results='hide', warning=FALSE, include=FALSE, eval=FALSE, cache=FALSE}
### Plotting themes

# -   The following plotting themes, colors, and typefaces are used throughout the project:

# define the plotting theme to be used in subsequent ggplots
luke_theme <-
  theme_bw() +
  theme(
    text = element_text(family = "Verdana"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    axis.title.x = element_text(size = 10),
    axis.text.x  = element_text(size = 8),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(linewidth = 0.5, colour = "grey40"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(linetype = "solid", colour = "grey"),
    legend.position.inside = c(0.1, 0.9)
  )
```

### custom functions

```{r}
# Function to extract month from date strings
extract_month <- function(date_str) {
  if (is.na(date_str)) {
    return(NA)
  }
  
  # Handle dates like "16/07/1887"
  if (grepl("^\\d{2}/\\d{2}/\\d{4}$", date_str)) {
    return(month(dmy(date_str)))
  }
  
  # Handle dates like "1974-07-26"
  if (grepl("^\\d{4}-\\d{1,2}-\\d{1,2}$", date_str)) {
    return(month(ymd(date_str)))
  }
  
  # Handle dates like "32118" (Excel date format)
  if (grepl("^\\d{4,5}$", date_str)) {
    return(month(as.Date(as.numeric(date_str), origin = "1899-12-30")))
  }
  
  # Handle 4-character year strings
  if (nchar(date_str) == 4) {
    return(NA)
  }
  
  return(NA)
}

# Function to extract year from date strings (handles Excel serials like "8753" correctly)
extract_year <- function(date_str) {
  if (is.na(date_str)) return(NA_real_)
  date_str <- trimws(as.character(date_str))
  
  # If string is only digits (no letters, punctuation)
  if (grepl("^\\d+$", date_str)) {
    num <- as.numeric(date_str)
    # If it's a plausible 4-digit calendar year, return it directly
    if (num >= 1800 && num <= 2100) {
      return(as.numeric(num))
    }
    # Otherwise treat as Excel serial date number and convert
    if (num >= 0 && num < 100000) {
      dt <- suppressWarnings(as.Date(num, origin = "1899-12-30"))
      if (!is.na(dt)) return(year(dt))
    }
    # fallback
    return(NA_real_)
  }
  
  # Full ISO-like date "1881-06-28" or "1894-8-30"
  if (grepl("^\\d{4}-\\d{1,2}-\\d{1,2}$", date_str)) {
    parsed <- suppressWarnings(ymd(date_str))
    if (!is.na(parsed)) return(year(parsed))
  }
  
  # DMY numeric slashes "16/07/1887"
  if (grepl("^\\d{1,2}/\\d{1,2}/\\d{4}$", date_str)) {
    parsed <- suppressWarnings(dmy(date_str))
    if (!is.na(parsed)) return(year(parsed))
  }
  
  # Try parsing day-month-year text variants, e.g. "9 May 1899", "26 January, 1892", "19, August, 1897"
  parsed <- suppressWarnings(dmy(gsub(",", "", date_str)))
  if (!is.na(parsed)) return(year(parsed))
  
  # Month-year formats like "Jul 1884", "July, 1874", "Aug 1881"
  if (grepl("[A-Za-z]+[^0-9A-Za-z]*\\d{4}", date_str)) {
    parsed2 <- suppressWarnings(dmy(paste0("01 ", gsub(",", "", date_str))))
    if (!is.na(parsed2)) return(year(parsed2))
  }
  
  # "pre-1944" -> extract 1944 (case-insensitive)
  if (grepl("^pre-\\d{4}$", date_str, ignore.case = TRUE)) {
    return(as.numeric(sub("(?i)pre-", "", date_str, perl = TRUE)))
  }
  
  # If there's a standalone 4-digit year anywhere (e.g. "1929 (likely)" or "(1929)")
  if (grepl("\\b\\d{4}\\b", date_str)) {
    yr <- regmatches(date_str, regexpr("\\b\\d{4}\\b", date_str))
    yr_num <- as.numeric(yr)
    # sanity-check year range
    if (!is.na(yr_num) && yr_num >= 1000 && yr_num <= 2100) return(yr_num)
  }
  
  # If nothing matched, return NA
  return(NA_real_)
}
```

# Triplicate analyses

This was conducted to assess the within-individual variation in feathers - we ideally want to be able to homogenize all breast feathers collected to increase the sample material but this assumes that there is limited within-individual variation among breast feathers.

For specimens that had enough material, we took 3 breast feathers and analyzed C, N, and O for each feather. The following analysis assesses the within- vs. among-individual variation in these three isotopes to provide insights on our ability to homogenize the feathers and use these isotopes for spatial inference.

### import and wrangle isotope data

```{r}
# import and consolidate key columns
triplicate_CN_dat <-
  read_excel(here("data/old file versions/Master_SI_Feathers_LEH_Sept24_v1_SB_v3_sample wts added.xls"), 
                     sheet = "Sample results_C&N", 
                     col_types = "text", skip = 10, col_names = TRUE) %>% 
  select(`Index #`, `Identifier 1`, `Amount`, `Ampl  Flash TCD...9`, `All Correct Amt % N Flash TCD`, `Ampl  28`, `normalised d15N`, `avg  Amt% N`, `stdev  Amt% N`, `avg  d15N`, `stdev d15N`, `Ampl  Flash TCD...18`, `All Correct Amt%C Flash TCD`, `Ampl  44`, `normalised d13C`, `Sample Dilution`, `avg  Amt% C`, `stdev  Amt% C`, `avg  d13C`, `stdev d13C`, `C:N Mass ratio`) %>% 
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>%
  mutate(
    feather = str_sub(identifier_1, -1),
    NewColumn = str_extract(identifier_1, "^[^ ]+"),  # Extract text to the left of the first white space
    identifier_1 = str_trim(str_replace(identifier_1, "^[^ ]+ ", ""))  # Remove text to the left of the first white space and trim remaining white space
  ) %>% mutate(identifier_1 = str_replace(identifier_1, "_.*", "")) %>% 
  rename(Specimen_ID = NewColumn,
         Feather_type = identifier_1) %>% 
  select(Specimen_ID, Feather_type, feather, amount, normalised_d15n, normalised_d13c) %>% 
  mutate(amount = as.numeric(amount), 
         normalised_d15n = as.numeric(normalised_d15n), 
         normalised_d13c = as.numeric(normalised_d13c)) %>%
  pivot_longer(
    cols = c(normalised_d15n, normalised_d13c),
    names_to = "isotope",
    values_to = "value"
  )

# import and consolidate key columns
triplicate_H_dat <-
  read_excel(here("data/Master_TCEA_H_SI_Batch 1_triplicate breast feathers_LEH_Nov25_SB.xls"), 
                     sheet = "Sample results_H", 
                     col_types = "text", skip = 9, col_names = TRUE) %>%
  slice(-11) %>% 
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>%
  dplyr::select(identifier_1, normalised_d2h) %>% 
  mutate(
    feather = str_extract(identifier_1, "(?<=_)."), 
    Feather_type = "Breast",
    Specimen_ID  = str_extract(identifier_1, "^[^_]+")
  ) %>% 
  dplyr::select(Specimen_ID, Feather_type, feather, normalised_d2h) %>% 
  mutate(normalised_d2h = as.numeric(normalised_d2h)) %>%
  pivot_longer(
    cols = c(normalised_d2h),
    names_to = "isotope",
    values_to = "value"
  )

# import and consolidate key columns
triplicate_O_dat <-
  read_excel(here("data/Master_TCEA_O_SI_Breast Feathers_Replication_Batch 4_LEH_Sept25_SB.xls"), 
                     sheet = "Sample results_O", 
                     col_types = "text", skip = 6, col_names = TRUE) %>%
  slice(-1) %>%
  slice(1:15) %>% 
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>%
  select(identifier_1, `amount`, 
         ampl_28, normalised_d18o) %>% 
    filter(!is.na(normalised_d18o)) %>%
  filter(identifier_1 != "AV21986_A feather") %>% 
  mutate(
    feather = str_extract(identifier_1, "(?<=_)."), 
    Feather_type = "Breast",
    Specimen_ID  = str_extract(identifier_1, "^[^_]+")
  ) %>% 
  select(Specimen_ID, Feather_type, feather, amount, normalised_d18o) %>% 
  mutate(amount = as.numeric(amount), 
         normalised_d18o = as.numeric(normalised_d18o)) %>%
  pivot_longer(
    cols = c(normalised_d18o),
    names_to = "isotope",
    values_to = "value"
  )

triplicate_CNOH_dat <- 
  bind_rows(triplicate_O_dat, triplicate_CN_dat, triplicate_H_dat)
```

### import and wrangle sample meta data

```{r}
# import meta data for all museum samples and add spatial data.
meta_data <-
  read_excel(here("data/MASTER_museum_ALL SAMPLES INVENTORY_banded_dotterels_v4.xlsx"), 
             col_types = "text") %>% 
  mutate(Specimen_ID_ = str_remove(Specimen_ID, regex(" (Breast|Primary)$", ignore_case = TRUE))) %>%
  mutate(Specimen_ID_ = str_remove_all(Specimen_ID_, regex("AJB ", ignore_case = TRUE)))  %>%
  mutate(Specimen_ID_ = str_remove_all(Specimen_ID_, regex(" -", ignore_case = TRUE)))%>% 
  select(Feather_type, Specimen_ID, Specimen_ID_, General_collection_location, 
         Specific_collection_location, Collection_date) %>% 
  mutate(Month = str_extract(Collection_date, "(Jan|Feb|Mar|Apr|May|Jun|Jul|Aug|Sep|Oct|Nov|Dec)")) %>% 
  mutate(Month = match(tolower(Month), tolower(month.abb))) %>%
  mutate(
    month_extracted = sapply(Collection_date, extract_month),
    Month = ifelse(is.na(Month) & !is.na(month_extracted), month_extracted, Month),
    Year = sapply(Collection_date, extract_year)
  ) %>% 
  select(-month_extracted) %>% 
  mutate(location = paste(Specific_collection_location, General_collection_location, sep = ", ")) %>%
  mutate(location = str_replace_all(location, "NA, ", "")) %>% 
  filter(!is.na(Specimen_ID))

# Create a data frame 
df <- data.frame(location = meta_data$location)

# Remove duplicates to avoid redundant API calls
unique_locations <- unique(df$location)

# Get latitude and longitude for unique locations
geocode_data <- geocode(unique_locations)

# Combine the geocode data with the unique locations
location_data <- data.frame(location = unique_locations, geocode_data) %>% filter(!is.na(location))

meta_data_ <-
  left_join(meta_data, location_data, by = "location") %>% 
  select(-c(Feather_type, Specimen_ID)) %>%
  distinct() %>% 
  rename(Specimen_ID = Specimen_ID_)
```

### merge triplicate data with metadata

```{r}
triplicate_CNOH_dat_meta <-
  stringdist_left_join(
    triplicate_CNOH_dat, meta_data_,
    by = "Specimen_ID",
    method = "cosine", # Jaro-Winkler distance
    # max_dist = 1,         # tweak based on how much difference you allow
    distance_col = "dist"
  ) %>% group_by(Specimen_ID.x, Feather_type, feather, isotope) %>% arrange(dist) %>% slice(1) %>%
  select(Specimen_ID.x, Specimen_ID.y, dist, everything()) %>% ungroup() %>% arrange(desc(dist)) %>% 
  select(-c(Specimen_ID.y, dist)) %>% 
  rename(Specimen_ID = Specimen_ID.x) %>% 
  mutate(geo_region = ifelse(str_detect(location, "Australia"), "Australia", 
                             ifelse(str_detect(location, "New Zealand"), "New Zealand", NA))) %>% 
  mutate(geo_region_bin = ifelse(geo_region == "Australia", 1, 0)) %>%
  mutate(
    month_name = month.name[Month],          # full month names
    month_abbrev = month.abb[Month],         # abbreviated names
    month_factor = factor(Month, levels = 1:12, labels = month.name, ordered = TRUE) # ordered factor
  ) %>% 
  mutate(lon = ifelse(lon < 0, lon + 180*2, lon)) %>% 
  mutate(Specific_collection_location = 
           ifelse(is.na(Specific_collection_location), General_collection_location, Specific_collection_location))
```

### calculate hulls for triplicates
C & N hulls
```{r}
# Compute convex hulls for each Specimen_ID using chull() from sp
hulls_triplicates_CN <- triplicate_CNOH_dat_meta %>%
  filter(!(isotope %in% c("normalised_d18o", "normalised_d2h"))) %>% 
   pivot_wider(
     names_from = isotope,
     values_from = value
   ) %>%
  group_by(Specimen_ID) %>%
  do({
    # Extract the data for the current group
    group_data <- .[ , c("normalised_d13c", "normalised_d15n")]
    
    # Create a spatial object but extract coordinates as a regular matrix
    coords <- SpatialPoints(group_data)
    coords_matrix <- coords@coords
    
    # Compute the convex hull using chull
    chull_indices <- chull(coords_matrix)
    hull_coords <- group_data[chull_indices, ]
    
    # Return the convex hull coordinates
    data.frame(hull = I(list(hull_coords)))
  }) %>%
  unnest(cols = c(hull)) %>% 
  left_join(., triplicate_CNOH_dat_meta %>% 
              dplyr::select(Specimen_ID, Month, location, lon, lat, geo_region, geo_region_bin) %>% distinct(), 
            by = "Specimen_ID")
```

#### among- vs. within-individual variation
```{r}
# install.packages("randomForest")

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(broom.mixed)
library(caret)
library(MASS)
library(randomForest)
library(patchwork)
library(brms)

# Clean and coerce types
dat <- triplicate_CNOH_dat_meta %>%
mutate(
Specimen_ID = as.factor(Specimen_ID),
Specific_collection_location = as.factor(Specific_collection_location),
feather = as.factor(feather),
isotope = as.character(isotope),
value = as.numeric(value)
) %>%
filter(!is.na(value)) %>%
  group_by(isotope) %>%
  mutate(value_scaled = as.numeric(scale(value))) %>%
  ungroup()  %>% 
  mutate(
    Month = factor(Month, levels = 1:12, labels = month.abb)
  )

CNOH_dat_meta_full <- readRDS(file = "data/CNOH_museum_samples_201125.rds")
CNOH_dat_meta_full %>% 
  group_by(Specimen_ID, isotope, Feather_type, value) %>% 
  summarise(mean_value = mean(value)) %>% 
  group_by(isotope) %>% 
  summarise(mean_iso_value = mean(mean_value),
            sd_iso_value = sd(mean_value))

triplicate_ids <- 
  CNOH_dat_meta_full %>% 
  filter(!is.na(feather)) %>% 
  pull(Specimen_ID) %>% 
  unique()

triplicate_dat <-
  CNOH_dat_meta_full %>% 
  filter(Specimen_ID %in% triplicate_ids) %>% 
  mutate(value_scaled = ifelse(isotope == "normalised_d13c", (value - -18.8) / 3.67,
                        ifelse(isotope == "normalised_d15n", (value - 10.9) / 2.63,
                        ifelse(isotope == "normalised_d18o", (value - 14.1) / 1.39,
                        ifelse(isotope == "normalised_d2h", (value - -21.6) / 22.0, NA))))) %>%
  mutate(
      Specimen_ID = as.factor(Specimen_ID),
      Specific_collection_location = as.factor(Specific_collection_location),
      feather = as.factor(feather),
      isotope = as.character(isotope),
      value_scaled = as.numeric(value_scaled)
    ) %>%
  filter(!is.na(value_scaled)) %>%
  ungroup()  %>% 
  mutate(
    Month = factor(Month, levels = 1:12, labels = month.abb)
  ) #%>% 
  # pivot_wider(
  #   names_from = isotope,
  #   values_from = value_scaled
  # ) %>% 
  # dplyr::select(Specimen_ID, Feather_type, region, normalised_d13c, normalised_d15n) %>% 
  #   filter(region != "homogenate") %>%
  # group_by(Specimen_ID, Feather_type, region) %>%
  # summarise(
  #   normalised_d13c = coalesce(normalised_d13c[!is.na(normalised_d13c)], NA_real_),
  #   normalised_d15n = coalesce(normalised_d15n[!is.na(normalised_d15n)], NA_real_),
  #   .groups = "drop"
  # )


ggplot() +
  geom_bar(
    data = triplicate_dat %>% dplyr::select(Specimen_ID, Month) %>% ungroup() %>% distinct(),
    aes(x = Month)
  ) +
  scale_x_discrete(drop = FALSE) +
  # facet_grid(geo_region ~ isotope) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  luke_theme

# --- Helper functions -------------------------------------------------------
get_varcomp_lmer <- function(mod) {
vc <- as.data.frame(VarCorr(mod)) %>% dplyr::select(grp, vcov)
names(vc) <- c("group", "variance")
total <- sum(vc$variance)
vc <- vc %>% mutate(prop = variance / total)
vc <- vc %>% arrange(desc(prop))
list(table = vc, total = total)
}


calc_repeatability <- function(mod, group = "Specimen_ID") {
vc <- as.data.frame(VarCorr(mod))
var_ind <- vc$vcov[vc$grp == group]
var_total <- sum(vc$vcov)
return(var_ind / var_total)
}

# --- Fit lmer models per isotope --------------------------------------------
lmer_results <- list()
vc_table <- tibble()
icc_table <- tibble()
isotopes <- unique(triplicate_dat$isotope)

for (iso in isotopes) {
message("Fitting lmer for ", iso)
d_iso <- triplicate_dat %>% filter(isotope == iso)


# Ensure there are repeated measures; if not, skip
if (n_distinct(d_iso$Specimen_ID) < 2) {
warning("Not enough individuals for isotope ", iso, " - skipping")
next
}


# Fit model
form <- as.formula("value_scaled ~ 1 + (1 | Specific_collection_location) + (1 | Specimen_ID)")
m <- tryCatch(lmer(form, data = d_iso, REML = TRUE), error = function(e) e)
if (inherits(m, "error")) {
warning("lmer failed for ", iso, ": ", m$message)
next
}


lmer_results[[iso]] <- m
vc <- get_varcomp_lmer(m)$table %>% mutate(isotope = iso)
vc_table <- bind_rows(vc_table, vc)
icc <- calc_repeatability(m, group = "Specimen_ID")
icc_table <- bind_rows(icc_table, tibble(isotope = iso, repeatability = icc))
}


# Tidy vc_table for plotting: ensure four groups present per isotope
vc_plot_df <- vc_table %>%
mutate(group = factor(group, levels = unique(group))) %>%
rename(variance = variance, proportion = prop)

# Tidy vc_table for plotting: ensure four groups present per isotope
vc_plot_df <- vc_table %>%
mutate(group = factor(group, levels = unique(group))) %>%
rename(variance = variance, proportion = prop)


# --- Plot variance partitioning per isotope ---------------------------------
# Prepare a plot dataframe from lmer results (preferred for clarity)
plot_df <- vc_table %>%
rename(component = group) %>%
mutate(component = recode(component,
"Specific_collection_location" = "Population",
"Specimen_ID" = "Individual",
"Residual" = "Residual")) %>%
group_by(isotope) %>%
mutate(component = factor(component, levels = c("Population","Individual","Feather","Residual"))) %>%
ungroup()


# p <- ggplot(plot_df, aes(x = isotope, y = prop, fill = component)) +
#   geom_col(position = "stack") +
#   scale_x_discrete(labels = iso_labels) +
#   labs(
#     x = "Isotope",
#     y = "Proportion of total variance",
#     fill = "Component",
#     title = "Variance partitioning per isotope"
#   ) +
#   theme_minimal() +
#   scale_fill_brewer(palette = "Pastel1")
# p

# Create a named vector mapping original isotope names to formatted labels
iso_labels <- c(
  "normalised_d13c" = "delta^{13}*C",
  "normalised_d15n" = "delta^{15}*N",
  "normalised_d2h"  = "delta^{2}*H",
  "normalised_d18o" = "delta^{18}*O"
)

vc_pie_chart <- 
  ggplot(plot_df, aes(x = "", y = prop, fill = component)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    facet_grid(isotope ~ ., labeller = labeller(isotope = iso_labels)) +
    labs(
        x = NULL,
        y = NULL,
        fill = "Variance\ncomponent\npartitioning",
        title = "Variance partitioning per isotope"
    ) +
    scale_fill_manual(values = c("grey20", "grey70", "grey90"),
    guide = guide_legend(
      title.position = "top",    # move title above keys
      title.hjust = 0.5,         # center title
      label.position = "right"   # labels stay to the right of keys
    )) +
    theme_minimal() +
    theme(
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        title = element_blank(),
        
        # Legend customization
        legend.position = "right",        # move legend to the right
        legend.direction = "vertical",    # legend items stacked (default), see below
        # legend.text = element_text(angle = 90)
        # legend.title = element_text(face = "bold", angle = 90),
        # legend.key.height = unit(0.5, "cm"),   # increase spacing between keys
        # legend.key.width  = unit(0.5, "cm"),
    
    # Increase white gap between keys
    # legend.spacing.x = unit(5, "cm") 
    )

vc_pie_chart

# Reorder individuals by population
dat_plot <- triplicate_dat %>%
  mutate(
    site = case_when(
      Specific_collection_location == "Waitangi West" ~ "Chatham Is.",
      Specific_collection_location == "Paraparaumu, Wellington" ~ "Wellington",
      Specific_collection_location == "Enderby Island"  ~ "Auckland Is.",
      Specific_collection_location == "Lake Ellesmere, Canterbury" ~ "Lake Ellesmere",
      Specific_collection_location == "Stewart Island, New Zealand"  ~ "Stewart Is.",
      TRUE ~ Specific_collection_location
    ),
    site = factor(site, 
                  levels = c("Auckland Is.", 
                             "Stewart Is.",
                             "Lake Ellesmere", 
                             "Chatham Is.",
                             "Wellington",
                             "Porirua")),
    Specimen_ID = fct_reorder(Specimen_ID, 
                              as.numeric(site), 
                              .fun = min),
    isotope = factor(isotope)
  )



scatterplot <- 
  ggplot(dat_plot, aes(
    x = Specimen_ID,
    y = value_scaled,
    colour = site
  )) +
  geom_jitter(
    size = 2.5,
    alpha = 0.8, width = 0.15
  ) +

  # Mean ± 95% CI per individual
  # stat_summary(
  #   fun = mean,
  #   fun.min = function(z) mean(z) - 1.96 * sd(z) / sqrt(length(z)),
  #   fun.max = function(z) mean(z) + 1.96 * sd(z) / sqrt(length(z)),
  #   geom = "pointrange",
  #   colour = "grey20",
  #   size = 0.4,
  #   position = position_nudge(x = 0.15)  # CI nudged right
  # ) +

  facet_grid(isotope ~ ., scales = "free_y",
             labeller = labeller(isotope = as_labeller(iso_labels, label_parsed))) +
  labs(
    x = "Individual",
    y = "Scaled isotope value",
    colour = "Population"
  ) +
  theme_bw(base_size = 13) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines"),
    legend.position = "none",
    
    # Rotate vertical strip text
    strip.text.y = element_text(angle = 0, size = 15),  # 0 = horizontal, 90 = vertical
    strip.background = element_blank()
  ) +
  scale_y_continuous(limits = c(-2.5, 2.5))

# Convert your table to an sf object
locations_sf <- dat_plot %>%
  dplyr::select(site, lon, lat) %>%
  distinct() %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326)  # WGS84

# fix point longitudes to -180..180
locations_fixed <- dat_plot %>%
  dplyr::select(site, lon, lat) %>%
  distinct() %>%
  mutate(
    lon = as.numeric(lon),
    lon = ifelse(lon > 180, lon - 360, lon)      # wrap >180 back into -180..180
  ) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# Get New Zealand map
nz <- ne_countries(scale = "medium", country = "New Zealand", returnclass = "sf")

library(ggrepel)

# Plot
nz_map <- 
  ggplot() +
  geom_sf(data = nz, fill = "grey95", color = "black") +
  geom_sf(data = locations_sf, aes(color = site), size = 3) +
  geom_text_repel(
    data = locations_sf,
    aes(x = st_coordinates(geometry)[,1],
        y = st_coordinates(geometry)[,2],
        label = site),
    size = 3,
    nudge_y = 0.5,
    max.overlaps = 50
  ) +
  coord_sf(xlim = c(165, 185), ylim = c(-51, -30)) +
  theme_minimal(base_size = 13) +
  labs(color = "Population", 
       title = "Sampling locations in New Zealand") +
  theme(
    axis.text = element_blank(), 
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    title = element_blank()
  )

nz_map + scatterplot + vc_pie_chart +
  plot_layout(widths = c(2, 2, 1), heights = c(3, 1, 1))
```

```{r}
dat_ <- 
  CNOH_dat_meta_full %>%
  ungroup() %>% 
  filter(Feather_type == "Breast") %>% 
  mutate(
    Specimen_ID = fct_reorder(as.factor(Specimen_ID), lon),
    Specific_collection_location = as.factor(Specific_collection_location),
    isotope = as.character(isotope),
    mean_value = as.numeric(mean_value)
  ) %>%
  filter(!is.na(mean_value)) %>%
  group_by(isotope) %>%
  mutate(value_scaled = as.numeric(scale(mean_value))) %>%
  ungroup() %>% 
  group_by(isotope, Specific_collection_location) %>% 
  mutate(n_inds = n_distinct(Specimen_ID)) #%>% 
  # filter(n_inds > 1)

dat_ %>% 
  filter(geo_region == "New Zealand" & isotope == "normalised_d2h") %>% 
  ggplot() +
  geom_point(aes(x = lon, y = value_scaled))

# --- Fit lmer models per isotope --------------------------------------------
lmer_results <- list()
vc_table <- tibble()
icc_table <- tibble()
isotopes <- unique(dat$isotope)

for (iso in isotopes) {
message("Fitting lmer for ", iso)
d_iso <- dat_ %>% filter(isotope == iso)


# Ensure there are repeated measures; if not, skip
if (n_distinct(d_iso$Specimen_ID) < 2) {
warning("Not enough individuals for isotope ", iso, " - skipping")
next
}


# Fit model
form <- as.formula("value_scaled ~ lat + lon + (1 | Specific_collection_location)")
m <- tryCatch(lmer(form, data = d_iso, REML = TRUE), error = function(e) e)
if (inherits(m, "error")) {
warning("lmer failed for ", iso, ": ", m$message)
next
}


lmer_results[[iso]] <- m
vc <- get_varcomp_lmer(m)$table %>% mutate(isotope = iso)
vc_table <- bind_rows(vc_table, vc)
icc <- calc_repeatability(m, group = "Specimen_ID")
icc_table <- bind_rows(icc_table, tibble(isotope = iso, repeatability = icc))
}


# Tidy vc_table for plotting: ensure four groups present per isotope
vc_plot_df <- vc_table %>%
mutate(group = factor(group, levels = unique(group))) %>%
rename(variance = variance, proportion = prop)

# Tidy vc_table for plotting: ensure four groups present per isotope
vc_plot_df <- vc_table %>%
mutate(group = factor(group, levels = unique(group))) %>%
rename(variance = variance, proportion = prop)


# --- Plot variance partitioning per isotope ---------------------------------
# Prepare a plot dataframe from lmer results (preferred for clarity)
plot_df <- vc_table %>%
rename(component = group) %>%
mutate(component = recode(component,
"Specific_collection_location" = "Population",
"Residual" = "Residual")) %>%
group_by(isotope) %>%
mutate(component = factor(component, levels = c("Population","Residual"))) %>%
ungroup()


# p <- ggplot(plot_df, aes(x = isotope, y = prop, fill = component)) +
#   geom_col(position = "stack") +
#   scale_x_discrete(labels = iso_labels) +
#   labs(
#     x = "Isotope",
#     y = "Proportion of total variance",
#     fill = "Component",
#     title = "Variance partitioning per isotope"
#   ) +
#   theme_minimal() +
#   scale_fill_brewer(palette = "Pastel1")
# p

vc_pie_chart <- 
  ggplot(plot_df, aes(x = "", y = prop, fill = component)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    facet_grid(isotope ~ ., labeller = labeller(isotope = iso_labels)) +
    labs(
        x = NULL,
        y = NULL,
        fill = "Variance\ncomponent\npartitioning",
        title = "Variance partitioning per isotope"
    ) +
    scale_fill_brewer(palette = "Pastel1",
    guide = guide_legend(
      title.position = "top",    # move title above keys
      title.hjust = 0.5,         # center title
      label.position = "right"   # labels stay to the right of keys
    )) +
    theme_minimal() +
    theme(
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        title = element_blank(),
        
        # Legend customization
        legend.position = "right",        # move legend to the right
        legend.direction = "vertical",    # legend items stacked (default), see below
        # legend.text = element_text(angle = 90)
        # legend.title = element_text(face = "bold", angle = 90),
        # legend.key.height = unit(0.5, "cm"),   # increase spacing between keys
        # legend.key.width  = unit(0.5, "cm"),
    
    # Increase white gap between keys
    # legend.spacing.x = unit(5, "cm") 
    )

vc_pie_chart

# Reorder individuals by population
dat_plot <- dat_ %>%
  mutate(
    Specimen_ID = fct_reorder(Specimen_ID, lat),
    isotope = factor(isotope)
  )

# Create a named vector mapping original isotope names to formatted labels
iso_labels <- c(
  "normalised_d13c" = "delta^{13}*C",
  "normalised_d15n" = "delta^{15}*N",
  "normalised_d2h"  = "delta^{2}*H",
  "normalised_d18o" = "delta^{18}*O"
)

scatterplot <- 
  ggplot(dat_plot, aes(
    x = Specimen_ID,
    y = value_scaled,
    colour = Specific_collection_location
  )) +
  geom_jitter(
    size = 2.5,
    alpha = 0.8, width = 0.15
  ) +

  # # Mean ± 95% CI per individual
  # stat_summary(
  #   fun = mean,
  #   fun.min = function(z) mean(z) - 1.96 * sd(z) / sqrt(length(z)),
  #   fun.max = function(z) mean(z) + 1.96 * sd(z) / sqrt(length(z)),
  #   geom = "pointrange",
  #   colour = "black",
  #   size = 0.4,
  #   position = position_nudge(x = 0.15)  # CI nudged right
  # ) +

  facet_grid(isotope ~ ., scales = "free_y",
             labeller = labeller(isotope = as_labeller(iso_labels, label_parsed))) +
  labs(
    x = "Individual",
    y = "Scaled isotope value",
    colour = "Population"
  ) +
  theme_bw(base_size = 13) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines"),
    legend.position = "none",
    
    # Rotate vertical strip text
    strip.text.y = element_text(angle = 0, size = 15),  # 0 = horizontal, 90 = vertical
    strip.background = element_blank()
  ) +
  geom_smooth()

scatterplot + vc_pie_chart

```


### Spatial d13C

```{r}
model_spat_C <- lmer(value ~ lat + lon + (1 | Specimen_ID), 
                     data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d13c"))
```

```{r}
spatial_effects <- allEffects(model_spat_C)
plot(spatial_effects)
```

```{r}
# Derive confidence intervals of effect sizes from parametric bootstrapping
tidy_spat_C <-
  tidy(model_spat_C, conf.int = TRUE, conf.method = "boot", nsim = 1000)

# Calculate marginal and conditional R-squared
r_squared <- r.squaredGLMM(model_spat_C)

# Display the results
print(r_squared)

# run rptR to obtain repeatabilities of random effects
rpt_spat_C <-
  rpt(value ~ lat + lon + (1 | Specimen_ID),
      grname = c("Specimen_ID", "Fixed"),
      data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d13c"),
      datatype = "Gaussian",
      nboot = 1000, npermut = 1000, ratio = TRUE,
      adjusted = TRUE, ncores = 4, parallel = TRUE)

# run partR2 on each model to obtain marginal R2, parameter estimates, and beta
# weights
R2m_spat_C <-
  partR2(model_spat_C,
         partvars = c("lat",
                      "lon"),
         R2_type = "marginal",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

R2c_spat_C <-
  partR2(model_spat_C,
         partvars = c("lat",
                      "lon"),
         R2_type = "conditional",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

stats_spat_C <-
  list(mod = model_spat_C,
       tidy = tidy_spat_C,
       rptR = rpt_spat_C,
       partR2m = R2m_spat_C,
       partR2c = R2c_spat_C,
       data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d13c"))

# summary(glht(smi_wing_lmer_seq))

#### Table of effect sizes ----
# Retrieve sample sizes
sample_sizes <-
  stats_spat_C$data %>%
  ungroup() %>%
  summarise(Individual = n_distinct(Specimen_ID),
            Observations = nrow(.))

sample_sizes <-
  as.data.frame(t(as.data.frame(sample_sizes))) %>%
  rownames_to_column("term") %>%
  rename(estimate = V1) %>%
  mutate(stat = "n")

# clean model component names
mod_comp_names <-
  data.frame(comp_name = c("baseline \U03B4 13 C (intercept)",
                           "latitude",
                           "longitude",
                           "Total Marginal \U1D479\U00B2",
                           "latitude",
                           "longitude",
                           "Total Conditional \U1D479\U00B2",
                           "Individual",
                           "Residual",
                           "Individual",
                           "Residual",
                           "Individuals",
                           "Observations"))

# Fixed effect sizes (non-standardized)
fixefTable <-
  stats_spat_C$tidy %>%
  dplyr::filter(effect == "fixed") %>%
  dplyr::select(term, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "fixed")

# Fixed effect sizes (standardized)
fixef_bw_Table <-
  stats_spat_C$partR2m$BW %>%
  as.data.frame() %>%
  mutate(stat = "fixed_bw") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Semi-partial R2 estimates
R2Table <-
  bind_rows(stats_spat_C$partR2m$R2,
            stats_spat_C$partR2c$R2[1,]) %>%
  dplyr::select(term, estimate, CI_lower, CI_upper) %>%
  as.data.frame() %>%
  mutate(stat = "partR2") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Random effects variances
ranefTable <-
  stats_spat_C$tidy %>%
  dplyr::filter(effect == "ran_pars") %>%
  dplyr::select(group, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "rand") %>%
  rename(term = group) %>%
  mutate(estimate = estimate^2,
         conf.high = conf.high^2,
         conf.low = conf.low^2)

# Adjusted repeatabilities
coefRptTable <-
  stats_spat_C$rptR$R_boot %>%
  dplyr::select(-Fixed) %>%
  mutate(residual = 1 - rowSums(.)) %>%
  apply(., 2,
        function(x) c(mean (x), quantile (x, prob = c(0.025, 0.975)))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(estimate = V1,
         conf.low = `2.5%`,
         conf.high = `97.5%`) %>%
  mutate(stat = "RptR")

# Store all parameters into a single table and clean it up
allCoefs_mod <-
  bind_rows(fixefTable[1,], 
            fixef_bw_Table,
            R2Table,
            ranefTable,
            coefRptTable,
            sample_sizes) %>%
  bind_cols(.,
            mod_comp_names) %>%
  mutate(coefString = ifelse(!is.na(conf.low),
                             paste0("[",
                                    round(conf.low, 2), ", ",
                                    round(conf.high, 2), "]"),
                             NA),
         effect = c(rep("Fixed effects \U1D6FD (standardized)", nrow(fixefTable[1,])),
                    rep("Fixed effects \U1D6FD (standardized)", nrow(fixef_bw_Table)),
                    rep("Partitioned \U1D479\U00B2", nrow(R2Table)),
                    rep("Random effects \U1D70E\U00B2", nrow(ranefTable)),
                    rep("Adjusted repeatability \U1D45F", nrow(coefRptTable)),
                    rep("Sample sizes \U1D45B", nrow(sample_sizes)))) %>%
  dplyr::select(effect, everything())

# draw gt table
stats_spat_C_table <-
  allCoefs_mod %>%
  dplyr::select(effect, comp_name, estimate, coefString) %>%
  gt(rowname_col = "row",
     groupname_col = "effect") %>%
  cols_label(comp_name = html("<i>\U03B4 13 C</i>"),
             estimate = "Mean estimate",
             coefString = "95% confidence interval") %>%
  fmt_number(columns = c(estimate),
             rows = 1:11,
             decimals = 2,
             use_seps = FALSE) %>%
  fmt_number(columns = c(estimate),
             rows = 12:13,
             decimals = 0,
             use_seps = FALSE) %>%
  sub_missing(columns = 1:4,
              missing_text = "") %>%
  cols_align(align = "left",
             columns = c(comp_name)) %>%
  tab_options(row_group.font.weight = "bold",
              row_group.background.color = brewer.pal(9,"Greys")[3],
              table.font.size = 12,
              data_row.padding = 3,
              row_group.padding = 4,
              summary_row.padding = 2,
              column_labels.font.size = 14,
              row_group.font.size = 12,
              table.width = pct(90))
```

```{r}
stats_spat_C_table
```

### Spatial d15N

```{r}
model_spat_N <- lmer(value ~ lat + lon + (1 | Specimen_ID), 
                     data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d15n"))
```

```{r}
spatial_effects <- allEffects(model_spat_N)
plot(spatial_effects)
```

```{r}
# Derive confidence intervals of effect sizes from parametric bootstrapping
tidy_spat_N<-
  tidy(model_spat_N, conf.int = TRUE, conf.method = "boot", nsim = 1000)

# run rptR to obtain repeatabilities of random effects
rpt_spat_N <-
  rpt(value ~ lat + lon + (1 | Specimen_ID),
      grname = c("Specimen_ID", "Fixed"),
      data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d15n"),
      datatype = "Gaussian",
      nboot = 1000, npermut = 1000, ratio = TRUE,
      adjusted = TRUE, ncores = 4, parallel = TRUE)

# run partR2 on each model to obtain marginal R2, parameter estimates, and beta
# weights
R2m_spat_N <-
  partR2(model_spat_N,
         partvars = c("lat",
                      "lon"),
         R2_type = "marginal",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

R2c_spat_N <-
  partR2(model_spat_N,
         partvars = c("lat",
                      "lon"),
         R2_type = "conditional",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

stats_spat_N <-
  list(mod = model_spat_N,
       tidy = tidy_spat_N,
       rptR = rpt_spat_N,
       partR2m = R2m_spat_N,
       partR2c = R2c_spat_N,
       data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d15n"))

# summary(glht(model_spat_N))

#### Table of effect sizes ----
# Retrieve sample sizes
sample_sizes <-
  stats_spat_N$data %>%
  ungroup() %>%
  summarise(Individual = n_distinct(Specimen_ID),
            Observations = nrow(.))

sample_sizes <-
  as.data.frame(t(as.data.frame(sample_sizes))) %>%
  rownames_to_column("term") %>%
  rename(estimate = V1) %>%
  mutate(stat = "n")

# clean model component names
mod_comp_names <-
  data.frame(comp_name = c("baseline \U03B4 15 N (intercept)",
                           "latitude",
                           "longitude",
                           "Total Marginal \U1D479\U00B2",
                           "latitude",
                           "longitude",
                           "Total Conditional \U1D479\U00B2",
                           "Individual",
                           "Residual",
                           "Individual",
                           "Residual",
                           "Individuals",
                           "Observations"))

# Fixed effect sizes (non-standardized)
fixefTable <-
  stats_spat_N$tidy %>%
  dplyr::filter(effect == "fixed") %>%
  dplyr::select(term, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "fixed")

# Fixed effect sizes (standardized)
fixef_bw_Table <-
  stats_spat_N$partR2m$BW %>%
  as.data.frame() %>%
  mutate(stat = "fixed_bw") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Semi-partial R2 estimates
R2Table <-
  bind_rows(stats_spat_N$partR2m$R2,
            stats_spat_N$partR2c$R2[1,]) %>%
  dplyr::select(term, estimate, CI_lower, CI_upper) %>%
  as.data.frame() %>%
  mutate(stat = "partR2") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Random effects variances
ranefTable <-
  stats_spat_N$tidy %>%
  dplyr::filter(effect == "ran_pars") %>%
  dplyr::select(group, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "rand") %>%
  rename(term = group) %>%
  mutate(estimate = estimate^2,
         conf.high = conf.high^2,
         conf.low = conf.low^2)

# Adjusted repeatabilities
coefRptTable <-
  stats_spat_N$rptR$R_boot %>%
  dplyr::select(-Fixed) %>%
  mutate(residual = 1 - rowSums(.)) %>%
  apply(., 2,
        function(x) c(mean (x), quantile (x, prob = c(0.025, 0.975)))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(estimate = V1,
         conf.low = `2.5%`,
         conf.high = `97.5%`) %>%
  mutate(stat = "RptR")

# Store all parameters into a single table and clean it up
allCoefs_mod <-
  bind_rows(fixefTable[1,], 
            fixef_bw_Table,
            R2Table,
            ranefTable,
            coefRptTable,
            sample_sizes) %>%
  bind_cols(.,
            mod_comp_names) %>%
  mutate(coefString = ifelse(!is.na(conf.low),
                             paste0("[",
                                    round(conf.low, 2), ", ",
                                    round(conf.high, 2), "]"),
                             NA),
         effect = c(rep("Fixed effects \U1D6FD (standardized)", nrow(fixefTable[1,])),
                    rep("Fixed effects \U1D6FD (standardized)", nrow(fixef_bw_Table)),
                    rep("Partitioned \U1D479\U00B2", nrow(R2Table)),
                    rep("Random effects \U1D70E\U00B2", nrow(ranefTable)),
                    rep("Adjusted repeatability \U1D45F", nrow(coefRptTable)),
                    rep("Sample sizes \U1D45B", nrow(sample_sizes)))) %>%
  dplyr::select(effect, everything())

# draw gt table
stats_spat_N_table <-
  allCoefs_mod %>%
  dplyr::select(effect, comp_name, estimate, coefString) %>%
  gt(rowname_col = "row",
     groupname_col = "effect") %>%
  cols_label(comp_name = html("<i>\U03B4 15 N</i>"),
             estimate = "Mean estimate",
             coefString = "95% confidence interval") %>%
  fmt_number(columns = c(estimate),
             rows = 1:11,
             decimals = 2,
             use_seps = FALSE) %>%
  fmt_number(columns = c(estimate),
             rows = 12:13,
             decimals = 0,
             use_seps = FALSE) %>%
  sub_missing(columns = 1:4,
              missing_text = "") %>%
  cols_align(align = "left",
             columns = c(comp_name)) %>%
  tab_options(row_group.font.weight = "bold",
              row_group.background.color = brewer.pal(9,"Greys")[3],
              table.font.size = 12,
              data_row.padding = 3,
              row_group.padding = 4,
              summary_row.padding = 2,
              column_labels.font.size = 14,
              row_group.font.size = 12,
              table.width = pct(90))
```

```{r}
stats_spat_N_table
```

### Spatial C:N

```{r, include=FALSE, eval=FALSE}
model_spat_CN <- lmer(value ~ lat + lon + (1 | Specimen_ID), 
                      data = triplicate_CNOH_dat_meta %>% filter(isotope == "cn_mass_ratio") %>% filter(!is.na(value) & !is.na(lat) & !is.na(lon) & !is.na(Specimen_ID)))

spatial_effects <- allEffects(model_spat_CN)
plot(spatial_effects)

# Derive confidence intervals of effect sizes from parametric bootstrapping
tidy_spat_CN<-
  tidy(model_spat_CN, conf.int = TRUE, conf.method = "boot", nsim = 1000)

# run rptR to obtain repeatabilities of random effects
rpt_spat_CN <-
  rpt(value ~ lat + lon + (1 | Specimen_ID),
      grname = c("Specimen_ID", "Fixed"),
      data = triplicate_CNOH_dat_meta %>% filter(isotope == "cn_mass_ratio") %>% filter(!is.na(value)),
      datatype = "Gaussian",
      nboot = 1000, npermut = 1000, ratio = TRUE,
      adjusted = TRUE, ncores = 4, parallel = TRUE)

# run partR2 on each model to obtain marginal R2, parameter estimates, and beta
# weights
R2m_spat_CN <-
  partR2(model_spat_CN,
         partvars = c("lat",
                      "lon"),
         R2_type = "marginal",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

R2c_spat_CN <-
  partR2(model_spat_CN,
         partvars = c("lat",
                      "lon"),
         R2_type = "conditional",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

stats_spat_CN <-
  list(mod = model_spat_CN,
       tidy = tidy_spat_CN,
       rptR = rpt_spat_CN,
       partR2m = R2m_spat_CN,
       partR2c = R2c_spat_CN,
       data = triplicate_CNOH_dat_meta %>% filter(isotope == "cn_mass_ratio") %>% filter(!is.na(value)))

# summary(glht(model_spat_CN))

#### Table of effect sizes ----
# Retrieve sample sizes
sample_sizes <-
  stats_spat_CN$data %>%
  ungroup() %>%
  summarise(Individual = n_distinct(Specimen_ID),
            Observations = nrow(.))

sample_sizes <-
  as.data.frame(t(as.data.frame(sample_sizes))) %>%
  rownames_to_column("term") %>%
  rename(estimate = V1) %>%
  mutate(stat = "n")

# clean model component names
mod_comp_names <-
  data.frame(comp_name = c("baseline \U03B4 13 C : \U03B4 15 N (intercept)",
                           "latitude",
                           "longitude",
                           "Total Marginal \U1D479\U00B2",
                           "latitude",
                           "longitude",
                           "Total Conditional \U1D479\U00B2",
                           "Individual",
                           "Residual",
                           "Individual",
                           "Residual",
                           "Individuals",
                           "Observations"))

# Fixed effect sizes (non-standardized)
fixefTable <-
  stats_spat_CN$tidy %>%
  dplyr::filter(effect == "fixed") %>%
  dplyr::select(term, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "fixed")

# Fixed effect sizes (standardized)
fixef_bw_Table <-
  stats_spat_CN$partR2m$BW %>%
  as.data.frame() %>%
  mutate(stat = "fixed_bw") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Semi-partial R2 estimates
R2Table <-
  bind_rows(stats_spat_CN$partR2m$R2,
            stats_spat_CN$partR2c$R2[1,]) %>%
  dplyr::select(term, estimate, CI_lower, CI_upper) %>%
  as.data.frame() %>%
  mutate(stat = "partR2") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Random effects variances
ranefTable <-
  stats_spat_CN$tidy %>%
  dplyr::filter(effect == "ran_pars") %>%
  dplyr::select(group, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "rand") %>%
  rename(term = group) %>%
  mutate(estimate = estimate^2,
         conf.high = conf.high^2,
         conf.low = conf.low^2)

# Adjusted repeatabilities
coefRptTable <-
  stats_spat_CN$rptR$R_boot %>%
  dplyr::select(-Fixed) %>%
  mutate(residual = 1 - rowSums(.)) %>%
  apply(., 2,
        function(x) c(mean (x), quantile (x, prob = c(0.025, 0.975)))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(estimate = V1,
         conf.low = `2.5%`,
         conf.high = `97.5%`) %>%
  mutate(stat = "RptR")

# Store all parameters into a single table and clean it up
allCoefs_mod <-
  bind_rows(fixefTable[1,], 
            fixef_bw_Table,
            R2Table,
            ranefTable,
            coefRptTable,
            sample_sizes) %>%
  bind_cols(.,
            mod_comp_names) %>%
  mutate(coefString = ifelse(!is.na(conf.low),
                             paste0("[",
                                    round(conf.low, 2), ", ",
                                    round(conf.high, 2), "]"),
                             NA),
         effect = c(rep("Fixed effects \U1D6FD (standardized)", nrow(fixefTable[1,])),
                    rep("Fixed effects \U1D6FD (standardized)", nrow(fixef_bw_Table)),
                    rep("Partitioned \U1D479\U00B2", nrow(R2Table)),
                    rep("Random effects \U1D70E\U00B2", nrow(ranefTable)),
                    rep("Adjusted repeatability \U1D45F", nrow(coefRptTable)),
                    rep("Sample sizes \U1D45B", nrow(sample_sizes)))) %>%
  dplyr::select(effect, everything())

# draw gt table
stats_spat_CN_table <-
  allCoefs_mod %>%
  dplyr::select(effect, comp_name, estimate, coefString) %>%
  gt(rowname_col = "row",
     groupname_col = "effect") %>%
  cols_label(comp_name = html("<i>\U03B4 13 C : \U03B4 15 N</i>"),
             estimate = "Mean estimate",
             coefString = "95% confidence interval") %>%
  fmt_number(columns = c(estimate),
             rows = 1:11,
             decimals = 2,
             use_seps = FALSE) %>%
  fmt_number(columns = c(estimate),
             rows = 12:13,
             decimals = 0,
             use_seps = FALSE) %>%
  sub_missing(columns = 1:4,
              missing_text = "") %>%
  cols_align(align = "left",
             columns = c(comp_name)) %>%
  tab_options(row_group.font.weight = "bold",
              row_group.background.color = brewer.pal(9,"Greys")[3],
              table.font.size = 12,
              data_row.padding = 3,
              row_group.padding = 4,
              summary_row.padding = 2,
              column_labels.font.size = 14,
              row_group.font.size = 12,
              table.width = pct(90))

stats_spat_CN_table
```

### Spatial d18O

```{r}
model_spat_O <- lmer(value ~ lat + lon + (1 | Specimen_ID), 
                     data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d18o"),
                     control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
```

```{r}
spatial_effects <- allEffects(model_spat_O)
plot(spatial_effects)
```

```{r}
# Derive confidence intervals of effect sizes from parametric bootstrapping
tidy_spat_O <-
  tidy(model_spat_O, conf.int = TRUE, conf.method = "boot", nsim = 1000)

# run rptR to obtain repeatabilities of random effects
rpt_spat_O <-
  rpt(value ~ lat + lon + (1 | Specimen_ID),
      grname = c("Specimen_ID", "Fixed"),
      data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d18o"),
      datatype = "Gaussian",
      nboot = 1000, npermut = 1000, ratio = TRUE,
      adjusted = TRUE, ncores = 4, parallel = TRUE)

# run partR2 on each model to obtain marginal R2, parameter estimates, and beta
# weights
R2m_spat_O <-
  partR2(model_spat_O,
         partvars = c("lat",
                      "lon"),
         R2_type = "marginal",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

R2c_spat_O <-
  partR2(model_spat_O,
         partvars = c("lat",
                      "lon"),
         R2_type = "conditional",
         nboot = 1000,
         CI = 0.95,
         max_level = 1)

stats_spat_O <-
  list(mod = model_spat_O,
       tidy = tidy_spat_O,
       rptR = rpt_spat_O,
       partR2m = R2m_spat_O,
       partR2c = R2c_spat_O,
       data = triplicate_CNOH_dat_meta %>% filter(isotope == "normalised_d18o"))

# summary(glht(model_spat_N))

#### Table of effect sizes ----
# Retrieve sample sizes
sample_sizes <-
  stats_spat_O$data %>%
  ungroup() %>%
  summarise(Individual = n_distinct(Specimen_ID),
            Observations = nrow(.))

sample_sizes <-
  as.data.frame(t(as.data.frame(sample_sizes))) %>%
  rownames_to_column("term") %>%
  rename(estimate = V1) %>%
  mutate(stat = "n")

# clean model component names
mod_comp_names <-
  data.frame(comp_name = c("baseline \U03B4 18 O (intercept)",
                           "latitude",
                           "longitude",
                           "Total Marginal \U1D479\U00B2",
                           "latitude",
                           "longitude",
                           "Total Conditional \U1D479\U00B2",
                           "Individual",
                           "Residual",
                           "Individual",
                           "Residual",
                           "Individuals",
                           "Observations"))

# Fixed effect sizes (non-standardized)
fixefTable <-
  stats_spat_O$tidy %>%
  dplyr::filter(effect == "fixed") %>%
  dplyr::select(term, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "fixed")

# Fixed effect sizes (standardized)
fixef_bw_Table <-
  stats_spat_O$partR2m$BW %>%
  as.data.frame() %>%
  mutate(stat = "fixed_bw") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Semi-partial R2 estimates
R2Table <-
  bind_rows(stats_spat_O$partR2m$R2,
            stats_spat_O$partR2c$R2[1,]) %>%
  dplyr::select(term, estimate, CI_lower, CI_upper) %>%
  as.data.frame() %>%
  mutate(stat = "partR2") %>%
  rename(conf.low = CI_lower,
         conf.high = CI_upper)

# Random effects variances
ranefTable <-
  stats_spat_O$tidy %>%
  dplyr::filter(effect == "ran_pars") %>%
  dplyr::select(group, estimate, conf.low, conf.high) %>%
  as.data.frame() %>%
  mutate(stat = "rand") %>%
  rename(term = group) %>%
  mutate(estimate = estimate^2,
         conf.high = conf.high^2,
         conf.low = conf.low^2)

# Adjusted repeatabilities
coefRptTable <-
  stats_spat_O$rptR$R_boot %>%
  dplyr::select(-Fixed) %>%
  mutate(residual = 1 - rowSums(.)) %>%
  apply(., 2,
        function(x) c(mean (x), quantile (x, prob = c(0.025, 0.975)))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(estimate = V1,
         conf.low = `2.5%`,
         conf.high = `97.5%`) %>%
  mutate(stat = "RptR")

# Store all parameters into a single table and clean it up
allCoefs_mod <-
  bind_rows(fixefTable[1,], 
            fixef_bw_Table,
            R2Table,
            ranefTable,
            coefRptTable,
            sample_sizes) %>%
  bind_cols(.,
            mod_comp_names) %>%
  mutate(coefString = ifelse(!is.na(conf.low),
                             paste0("[",
                                    round(conf.low, 2), ", ",
                                    round(conf.high, 2), "]"),
                             NA),
         effect = c(rep("Fixed effects \U1D6FD (standardized)", nrow(fixefTable[1,])),
                    rep("Fixed effects \U1D6FD (standardized)", nrow(fixef_bw_Table)),
                    rep("Partitioned \U1D479\U00B2", nrow(R2Table)),
                    rep("Random effects \U1D70E\U00B2", nrow(ranefTable)),
                    rep("Adjusted repeatability \U1D45F", nrow(coefRptTable)),
                    rep("Sample sizes \U1D45B", nrow(sample_sizes)))) %>%
  dplyr::select(effect, everything())

# draw gt table
stats_spat_O_table <-
  allCoefs_mod %>%
  dplyr::select(effect, comp_name, estimate, coefString) %>%
  gt(rowname_col = "row",
     groupname_col = "effect") %>%
  cols_label(comp_name = html("<i>\U03B4 18 O</i>"),
             estimate = "Mean estimate",
             coefString = "95% confidence interval") %>%
  fmt_number(columns = c(estimate),
             rows = 1:11,
             decimals = 2,
             use_seps = FALSE) %>%
  fmt_number(columns = c(estimate),
             rows = 12:13,
             decimals = 0,
             use_seps = FALSE) %>%
  sub_missing(columns = 1:4,
              missing_text = "") %>%
  cols_align(align = "left",
             columns = c(comp_name)) %>%
  tab_options(row_group.font.weight = "bold",
              row_group.background.color = brewer.pal(9,"Greys")[3],
              table.font.size = 12,
              data_row.padding = 3,
              row_group.padding = 4,
              summary_row.padding = 2,
              column_labels.font.size = 14,
              row_group.font.size = 12,
              table.width = pct(90))
```

```{r}
stats_spat_O_table
```

## Different sampling and homogenization proceedures
half a vane vs. sub-sample of scissor cut vane homogenate

```{r}
# import and consolidate key columns
dat <-
  read_excel(here("data/Master_SI_Replicate test Feathers_N_9_LEH_Jan25_v1_SB.xls"), 
                     sheet = "Sample results_C&N", 
                     col_types = "text", skip = 9, col_names = TRUE) %>%
  dplyr::select(1, 2, 10:16, 18:24, 28:35, 37, 40:46, 48) %>%
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>% 
  rename(all_correct_amt_perc__n_flash_tcd_____A = all_correct_amt_perc__n_flash_tcd...10, 
         `ampl_28_mv_____A` = `ampl_28_(mv)...11`, 
         normalised_d15n_____A = normalised_d15n...12,                
         avg_amt_perc_n_____A = avg_amt_perc_n...13,             
         stdev_amt_perc_n_____A = stdev_amt_perc_n...14,                  
         avg_d15n_____A = avg_d15n...15,    
         stdev_d15n_____A = stdev_d15n...16,                        
         all_correct_amt_perc__n_flash_tcd_____B = all_correct_amt_perc__n_flash_tcd...18,
         `ampl_28_mv_____B` = `ampl_28_(mv)...19`,                     
         normalised_d15n_____B = normalised_d15n...20,    
         avg_amt_perc_n_____B = avg_amt_perc_n...21,                    
         stdev_amt_perc_n_____B = stdev_amt_perc_n...22,  
         avg_d15n_____B = avg_d15n...23,                          
         stdev_d15n_____B = stdev_d15n...24,   
         all_correct_amt_perc__c_flash_tcd_____C = all_correct_amt_perc__c_flash_tcd...28,
         ampl_44_mv_____C = `ampl_44_(mv)...29`,                      
         normalised_d13c_____C = normalised_d13c...30, 
         sample_dilution__perc_____C = `sample_dilution_(_perc_)`,               
         avg_amt_perc_c_____C = avg_amt_perc_c...32,   
         stdev_amt_perc_c_____C = stdev_amt_perc_c...33,                       
         avg_d13c_____C = avg_d13c...34, 
         stdev_d13c_____C = stdev_d13c...35,                             
         cn_mass_ratio_____C = `cn_mass_ratio,_second_batch`,            
         all_correct_amt_perc__c_flash_tcd_____D = all_correct_amt_perc__c_flash_tcd...40,
         ampl_44_mv_____D = `ampl_44_(mv)...41`,                      
         normalised_d13c_____D = normalised_d13c...42, 
         avg_amt_perc_C_____D = avg_amt_perc_c...43,                    
         stdev_amt_perc_C_____D = stdev_amt_perc_c...44,                 
         avg_d13C_____D = avg_d13c...45,                          
         stdev_d13C_____D = stdev_d13c...46,                       
         cn_mass_ratio_____D = `cn_mass_ratio,_first_batch`) %>% 
  pivot_longer(
    cols = -c(index_, identifier_1), 
    names_to = c("Group", "Variable"), 
    names_sep = "_____"
  ) %>% 
  mutate(method = ifelse(Variable %in% c("A", "C"), "B", "A"),
         element = ifelse(Variable %in% c("A", "B"), "N", "C"),
         value = as.numeric(value)) %>% 
  mutate(
    NewColumn = str_extract(identifier_1, "^[^ ]+"),  # Extract text to the left of the first white space
    feather_id = str_trim(str_replace(identifier_1, "^[^ ]+ ", "")),
    identifier_1 = str_trim(str_replace(identifier_1, "^[^ ]+ ", ""))# Remove text to the left of the first white space and trim remaining white space
  ) %>% mutate(identifier_1 = str_replace(identifier_1, "_.*", "")) %>% 
  rename(Specimen_ID = NewColumn,
         Feather_type = identifier_1) %>% 
  left_join(., read_excel(here("../../stable_isotopes/data/for_Sarah/master_museum_sample_banded_dotterels.xlsx"), col_types = "text") %>% mutate(Specimen_ID = str_extract(Specimen_ID, "^[^ ]+")) %>% dplyr::select(Feather_type, Specimen_ID, General_collection_location, 
         Specific_collection_location, Collection_date), join_by(Specimen_ID, Feather_type)) %>% 
  dplyr::select(Specimen_ID, feather_id, Feather_type, Group, method, 
         element, General_collection_location, 
         Specific_collection_location, Collection_date, value) %>%
  mutate(Month = str_extract(Collection_date, "(January|February|March|April|May|June|July|August|September|October|November|December)")) %>% 
  mutate(location = paste(Specific_collection_location, General_collection_location, sep = ", ")) %>%
  mutate(location = str_replace_all(location, "NA, ", ""))

dat_C <-
  dat %>% filter(element == "C") %>% 
  pivot_wider(names_from = Group, values_from = value)

dat_N <-  
  dat %>% filter(element == "N") %>% 
  pivot_wider(names_from = Group, values_from = value)
  
# Create a data frame 
df <- data.frame(location = dat$location)

# Remove duplicates to avoid redundant API calls
unique_locations <- unique(df$location)

# Get latitude and longitude for unique locations
geocode_data <- geocode(unique_locations)

# Combine the geocode data with the unique locations
location_data <- data.frame(location = unique_locations, geocode_data)

# Merge the geocode data back into the original dataframe
dat_C <- left_join(dat_C, location_data, by = "location")
dat_N <- left_join(dat_N, location_data, by = "location")

# transform lon
dat_C <-
  dat_C %>% 
  mutate(lon = ifelse(lon < 0, lon + 180*2, lon))

dat_N <-
  dat_N %>% 
  mutate(lon = ifelse(lon < 0, lon + 180*2, lon))

dat_C_mean <-
  dat_C %>%
  group_by(Specimen_ID, method) %>%
  summarise(normalised_d13c = mean(normalised_d13c))

dat_N_mean <-
  dat_N %>%
  group_by(Specimen_ID, method) %>%
  summarise(normalised_d15n = mean(normalised_d15n))

dat_NC_mean <- 
  left_join(dat_C_mean, dat_N_mean)

dat_N_ <- 
  dat_N %>% 
  dplyr::select(method, Specimen_ID, feather_id, normalised_d15n)

dat_C_ <- 
  dat_C %>% 
  dplyr::select(method, Specimen_ID, feather_id, normalised_d13c)

dat_NC_2 <- 
  left_join(dat_C_, dat_N_)

# Compute convex hulls for each Specimen_ID using chull() from sp
hulls_ <- dat_NC_2 %>%
  group_by(method, Specimen_ID) %>%
  do({
    # Extract the data for the current group
    group_data <- .[ , c("normalised_d13c", "normalised_d15n")]
    
    # Create a spatial object but extract coordinates as a regular matrix
    coords <- SpatialPoints(group_data)
    coords_matrix <- coords@coords
    
    # Compute the convex hull using chull
    chull_indices <- chull(coords_matrix)
    hull_coords <- group_data[chull_indices, ]
    
    # Return the convex hull coordinates
    data.frame(hull = I(list(hull_coords)))
  }) %>%
  unnest(cols = c(hull)) %>% 
  left_join(., dat %>% dplyr::select(Specimen_ID, Month, location) %>% distinct(), by = "Specimen_ID") %>% 
  mutate(Specimen_ID_method = paste(Specimen_ID, method, sep = "_"))

ggplot() + 
  geom_polygon(data = hulls_, 
               aes(group = Specimen_ID_method, fill = location, 
                   x = normalised_d13c, y = normalised_d15n), 
               color = NA, alpha = 0.5) +
  geom_point(data = hulls_, 
             aes(fill = location, x = normalised_d13c, y = normalised_d15n, shape = method), 
             size = 3) + 
  labs(x = "d13C", y = "d15N") + 
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  scale_shape_manual(values = c(21, 24), labels = c("A" = "homogenate of entire half vane",
                                                    "B" = "homogenate of vane sub-sample")) +
  theme(legend.position = "bottom", legend.title = element_blank(),
        axis.title = element_text(size = 15)) + 
  guides(color = guide_legend(nrow = 3), fill = guide_legend(nrow = 3), shape = guide_legend(nrow = 3)) +
  geom_point(data = dat_NC_mean, aes(x = normalised_d13c, y = normalised_d15n, shape = method), fill = "black")
```

## assess quantitative differences in the isotope data according to sampling/homogenization method

```{r, eval=FALSE}
C_amt_perc_mod <- 
  lmer(all_correct_amt_perc__c_flash_tcd ~ method + (1 | Specimen_ID), 
       data = dat_C)

double_model = bf(all_correct_amt_perc__c_flash_tcd ~ method + gr(Specimen_ID, by = method),
                  sigma ~ method)

m3_brm <- brm(double_model,
              data = dat_C,
              warmup = 500,iter = 3000, thin=2,
              chains = 2, inits = "random",
              seed = 12345)


C_ampl_44_mv_mod <- 
  lmer(ampl_44_mv ~ method + (1 | Specimen_ID), 
       data = dat_C)

C_normalised_d13c_mod <- 
  lmer(normalised_d13c ~ method + (1 | Specimen_ID), 
       data = dat_C)

N_amt_perc_mod <- 
  lmer(all_correct_amt_perc__n_flash_tcd ~ method + (1 | Specimen_ID), 
       data = dat_N)

N_ampl_28_mv_mod <- 
  lmer(ampl_28_mv ~ method + (1 | Specimen_ID), 
       data = dat_N)

N_normalised_d15n_mod <- 
  lmer(normalised_d15n ~ method + (1 | Specimen_ID), 
       data = dat_N)
```

## Sample management

```{r}
museum_samples <- 
  read_sheet("https://docs.google.com/spreadsheets/d/1-iF-rvswS5QvlKgz4fgF14FnfkK17DYf6je5nwYw48o/edit?gid=0#gid=0", col_types = "c") %>% 
  filter(Sent_to_Sarah == "yes" | Sent_to_Sarah == "Yes") %>%
  mutate(Month = str_extract(Collection_date, "(January|February|March|April|May|June|July|August|September|October|November|December)")) %>% 
  mutate(location = paste(gsub("\\s*\\([^\\)]+\\)", "", Specific_collection_location), General_collection_location, sep = ", ")) %>%
  mutate(location = str_replace_all(location, "NA, ", ""))

# Applying the transformation
museum_samples <- museum_samples %>%
  mutate(
    month_extracted = sapply(Collection_date, extract_month),
    month = ifelse(!is.na(Collection_month), as.numeric(Collection_month), month_extracted)
  ) %>%
  select(-month_extracted)

# Create a data frame 
museum_sample_loc <- 
  data.frame(location = museum_samples$location) %>% 
  pull(location) %>% unique()

geocode_data <- geocode(museum_sample_loc)

# Combine the geocode data with the unique locations
museum_sample_loc_ <- data.frame(location = museum_sample_loc, geocode_data)

# Merge the geocode data back into the original dataframe
museum_samples <- left_join(museum_samples, museum_sample_loc_, by = "location")

# Remove rows where month is NA
museum_samples <- museum_samples %>% filter(!is.na(month))

# Convert numeric month to month names
museum_samples$month <- factor(month.name[museum_samples$month], levels = month.name)

# Convert museum_samples to an sf object
museum_samples_sf <- st_as_sf(museum_samples, coords = c("lon", "lat"), crs = 4326)

# Reproject museum_samples to the desired projection
museum_samples_sf <- st_transform(museum_samples_sf, crs = "+proj=ortho +lon_0=160 +lat_0=-41")

museum_samples_sf <- 
  museum_samples_sf %>%
  mutate(Specimen_ID_ = gsub("WAM |AJB ", "", Specimen_ID)) %>%
  mutate(Feather_type = ifelse(is.na(Feather_type), sub("^(\\S+\\s+\\S+)\\s+(.*)$", "\\2", Specimen_ID), Feather_type)) %>% 
  mutate(
    Specimen_ID_ = str_extract(Specimen_ID_, "^[^ ]+")) %>%   # Extract text to the left of the first white space
  group_by(Specimen_ID_) %>%
  mutate(Feather_status = case_when(
    n_distinct(Feather_type) == 2 ~ "Primary and Breast",
    "Breast" %in% Feather_type ~ "Breast",
    "Primary" %in% Feather_type ~ "Primary"
  )) %>%
  ungroup()

museum_samples_sf %>% dplyr::select(Specimen_ID_, Feather_status) %>% distinct()

# Get world map data
world <- ne_countries(scale = "medium", returnclass = "sf")

# Extract the coordinates from the sf object
museum_samples_df <- 
  st_as_sf(museum_samples_sf %>% 
             dplyr::select(Museum, Specimen_ID_, Sex, location, Feather_status, month) %>% distinct()) %>%
  st_coordinates() %>%
  as.data.frame() %>%
  bind_cols(st_drop_geometry(museum_samples_sf %>% 
             dplyr::select(Museum, Specimen_ID_, Sex, location, Feather_status, month) %>% distinct()))

# Create the plot with projection centered on 180° longitude and custom bounds
ggplot() +
  geom_sf(data = world, fill = "lightgrey") +  # Add world map
  geom_point(data = museum_samples_df %>% 
               filter(month %in% c("May", "June", "July", "August")) %>% 
               mutate(geo_region = ifelse(str_detect(location, "Zealand"), "NZ", "AU")), 
             aes(x = X, y = Y, fill = geo_region),
             shape = 21, 
             size = 3, 
             alpha = 0.8, 
             position = position_jitter(width = 50000, height = 50000)) +  # Add jitter to points
  # facet_wrap(~ month, ncol = 3) +  # Create facets for each month
  coord_sf(xlim = c(-2312396, 1837687), ylim = c(-866762, 885204), 
           crs = "+proj=ortho +lon_0=160 +lat_0=-41") +  # Custom bounds for x and y axes
  theme_minimal() +  # Use a minimal theme
  labs(title = "Museum specimens collected in winter") +#, 
       # subtitle = "Collection Locations by Month", 
       # x = "Longitude", y = "Latitude") +
  theme(panel.grid.minor = element_line(linewidth = 0.5, colour = "grey40"),
        axis.title = element_blank(), legend.position = "none") +
  scale_fill_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9"))

museum_samples_df %>% 
  filter(str_detect(location, "Australia") | 
           (str_detect(location, "New Zealand") & 
              month %in% c("May", "June", "July"))) %>% 
  group_by(location, X, Y, month) %>%
  mutate(n_ind_at_location = n_distinct(Specimen_ID_)) %>% 
  ungroup() %>% 
  dplyr::select(-X, -Y) %>% 
  # filter(n_ind_at_location > 2) %>% View()
  arrange(desc(n_ind_at_location)) %>% 
  write.csv("data/museum_samples_to_analyze_161224.csv", row.names = FALSE)

museum_samples_df %>% 
  filter(str_detect(location, "Australia") | 
           (str_detect(location, "New Zealand") & 
              month %in% c("May", "June", "July"))) %>% 
  group_by(location, X, Y, month) %>%
  mutate(n_ind_at_location = n_distinct(Specimen_ID_)) %>% 
  ungroup() %>% 
  dplyr::select(-X, -Y) %>%
  nrow()
```

```{r}
museum_samples %>% 
  group_by(Museum) %>% 
  summarise(n_museums = n_distinct(Specimen_ID)) %>%
  bind_rows(data.frame(Museum = "Total", 
                       n_museums = museum_samples %>% 
                         summarise(x = n_distinct(Specimen_ID)) %>% 
                         pull(x))) %>% 
  gt() %>% 
  cols_label(Museum = "Museum",
             n_museums = "No. specimens") %>% 
  tab_options(row_group.font.weight = "bold",
              row_group.background.color = brewer.pal(9,"Greys")[3],
              table.font.size = 12,
              data_row.padding = 3,
              row_group.padding = 4,
              summary_row.padding = 2,
              column_labels.font.size = 14,
              row_group.font.size = 12,
              table.width = pct(90))
```

## checking feather barbs vs. feather shaft

```{r}
dat <-
  read_excel(here("data/Master_EA_C_N_SI_Batch 3 Breast feathers_shaft and barb comparison_LEH_Sept25_v1_SB.xls"), 
                     sheet = "Sample results_C&N", 
                     col_types = "text", skip = 9, col_names = TRUE) %>%
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>%
  rename(all_correct_amt_perc__n_flash_tcd = all_correct_amt_perc__n_flash_tcd, 
         `ampl_28_mv` = `ampl_28_(mv)`, 
         normalised_d15n = normalised_d15n,                
         avg_amt_perc_n = avg_amt_perc__n,             
         stdev_amt_perc_n = stdev_amt_perc__n,                  
         avg_d15n = avg_d15n,    
         stdev_d15n = stdev_d15n,                        
         all_correct_amt_perc__c_flash_tcd = all_correct_amt_perc__c_flash_tcd,
         ampl_44_mv = `ampl_44_(mv)`,
         normalised_d13c = normalised_d13c,
         sample_dilution__perc = `sample_dilution_(_perc_)`,
         avg_amt_perc_c = avg_amt_perc__c,
         stdev_amt_perc_c = stdev_amt_perc__c,
         avg_d13c = avg_d13c,
         stdev_d13c = stdev_d13c,
         cn_mass_ratio = `cn_mass_ratio`) %>% 
    filter(!is.na(index_)) %>%
  mutate(
    Feather_type = str_extract(identifier_1, regex("Primary|Breast", ignore_case = TRUE)),
    region       = str_extract(identifier_1, regex("Barb|Shaft", ignore_case = TRUE)),
    Specimen_ID  = str_extract(identifier_1, "^[^_]+")
  ) %>% 
  dplyr::select(Specimen_ID, Feather_type, region, c(4,9:35)) %>% 
  left_join(., read_excel(here("../../stable_isotopes/data/for_Sarah/master_museum_sample_banded_dotterels.xlsx"), col_types = "text") %>% mutate(Specimen_ID = str_extract(Specimen_ID, "^[^ ]+")) %>% dplyr::select(Feather_type, Specimen_ID, General_collection_location, 
         Specific_collection_location, Collection_date), join_by(Specimen_ID, Feather_type)) %>% 
  mutate(Month = str_extract(Collection_date, "(January|February|March|April|May|June|July|August|September|October|November|December)")) %>% 
  mutate(location = paste(Specific_collection_location, General_collection_location, sep = ", ")) %>%
  mutate(location = str_replace_all(location, "NA, ", "")) %>% 
  mutate(region = ifelse(region == "Shaft", "Rachis", region))

# Create a data frame 
df <- data.frame(location = dat$location)

# Remove duplicates to avoid redundant API calls
unique_locations <- unique(df$location)

# Get latitude and longitude for unique locations
geocode_data <- geocode(unique_locations)

# Combine the geocode data with the unique locations
location_data <- data.frame(location = unique_locations, geocode_data)

dat <- left_join(dat, location_data, by = "location")

CNOH_dat_meta_full <- readRDS(file = "data/CNOH_museum_samples_201125.rds")
CNOH_dat_meta_full %>% 
  group_by(Specimen_ID, isotope, Feather_type, value) %>% 
  summarise(mean_value = mean(value)) %>% 
  group_by(isotope) %>% 
  summarise(mean_iso_value = mean(mean_value),
            sd_iso_value = sd(mean_value))

barb_shaft_ids <- 
  CNOH_dat_meta_full %>% 
  filter(region != "homogenate") %>% 
  pull(Specimen_ID) %>% 
  unique()

barb_shaft_dat <-
  CNOH_dat_meta_full %>% 
  filter(Specimen_ID %in% barb_shaft_ids) %>% 
  mutate(value_scaled = ifelse(isotope == "normalised_d13c", (value - -18.8) / 3.67,
                        ifelse(isotope == "normalised_d15n", (value - 10.9) / 2.63,
                        ifelse(isotope == "normalised_d18o", (value - 14.1) / 1.39,
                        ifelse(isotope == "normalised_d2h", (value - -21.6) / 22.0, NA))))) %>% 
  pivot_wider(
    names_from = isotope,
    values_from = value_scaled
  ) %>% 
  dplyr::select(Specimen_ID, Feather_type, region, normalised_d13c, normalised_d15n) %>% 
    filter(region != "homogenate") %>%
  group_by(Specimen_ID, Feather_type, region) %>%
  summarise(
    normalised_d13c = coalesce(normalised_d13c[!is.na(normalised_d13c)], NA_real_),
    normalised_d15n = coalesce(normalised_d15n[!is.na(normalised_d15n)], NA_real_),
    .groups = "drop"
  )

ggplot() + 
  geom_line(data = barb_shaft_dat,
             aes(x = as.numeric(normalised_d13c), y = as.numeric(normalised_d15n), 
                 color = Feather_type, group = Specimen_ID), 
             size = 0.5, alpha = 0.5) + 
  geom_point(data = barb_shaft_dat,
             aes(x = as.numeric(normalised_d13c), y = as.numeric(normalised_d15n), 
                 color = Feather_type, shape = region), 
             size = 3) + 
    # geom_point(data = dat_C_N %>% select(Specimen_ID, avg_d13C_12C, avg_d15N_14N) %>% distinct(),
    #          aes(x = avg_d13C_12C, y = avg_d15N_14N, 
    #              color = Specimen_ID), shape = 3, 
    #          size = 6) + 
  labs(x = expression(delta^{13}*C[scaled]), y = expression(delta^{15}*N[scaled]), shape = "Region",
  color = "Feather") + 
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom", legend.title = element_blank(),
        axis.title = element_text(size = 15)) +
  luke_theme
```

## Batch 3 and 4 - C, N, & O

### import and wrangle C, N, O, and H data

```{r}
CN_dat <-
  read_excel(here("data/Master_EA_C_N_SI_Batch 3 vane_barb and Batch 4 Breast Feathers_LEH_Sept25_v1_SB.xls"), 
                     sheet = "Sample results_C&N", 
                     col_types = "text", skip = 9, col_names = TRUE) %>%
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>% 
  dplyr::select(`identifier_1`, `amount_(mg)`, 
         `ampl_flash_tcd_(mv)...10`, 
         all_correct_amt_perc__n_flash_tcd, 
         `ampl_28_(mv)`, 
         normalised_d15n, 
         `ampl_flash_tcd_(mv)...15`, 
         all_correct_amt_perc__c_flash_tcd, 
         `ampl_44_(mv)`, 
         normalised_d13c, 
         cn_mass_ratio) %>% 
    filter(!is.na(normalised_d15n)) %>%
  rename(amount = `amount_(mg)`) %>% 
  mutate(
    Feather_type = str_extract(identifier_1, regex("Primary|Breast", ignore_case = TRUE)),
    region       = str_extract(identifier_1, regex("Barb|Shaft", ignore_case = TRUE)),
    region       = if_else(is.na(region), "homogenate", region),
    Specimen_ID  = str_extract(identifier_1, "^[^_]+")
  ) %>%
  dplyr::select(Specimen_ID, Feather_type, region, amount, normalised_d15n, normalised_d13c) %>% 
  mutate(amount = as.numeric(amount), 
         normalised_d15n = as.numeric(normalised_d15n), 
         normalised_d13c = as.numeric(normalised_d13c)) %>%
  pivot_longer(
    cols = c(normalised_d15n, normalised_d13c),
    names_to = "isotope",
    values_to = "value"
  )

O_dat <-
  read_excel(here("data/Master_TCEA_O_SI_Breast Feathers_Replication_Batch 4_LEH_Sept25_SB.xls"), 
                     sheet = "Sample results_O", 
                     col_types = "text", skip = 6, col_names = TRUE) %>%
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>%
  dplyr::select(sia_id, `amount`, 
         ampl_28, normalised_d18o) %>% 
    filter(!is.na(normalised_d18o)) %>%
  slice(16:44) %>% 
  mutate(
    Feather_type = "Breast",
    region       = "homogenate") %>% 
  rename(Specimen_ID  = sia_id) %>% 
  dplyr::select(Specimen_ID, Feather_type, region, amount, normalised_d18o) %>% 
  mutate(amount = as.numeric(amount), 
         normalised_d18o = as.numeric(normalised_d18o)) %>%
  pivot_longer(
    cols = c(normalised_d18o),
    names_to = "isotope",
    values_to = "value"
  )

H_dat <-
  read_excel(here("data/Master_TCEA_H_SI_BATCH 4 BREAST Feathers_LE_Nov25_v1.xls"), 
                     sheet = "Sample results_H", 
                     col_types = "text", skip = 9, col_names = TRUE) %>%
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>%
  dplyr::select(sia_id, ampl_2, normalised_d2h) %>% 
    filter(!is.na(normalised_d2h)) %>% 
  mutate(
    Feather_type = "Breast",
    region       = "homogenate") %>% 
  rename(Specimen_ID  = sia_id) %>% 
  dplyr::select(Specimen_ID, Feather_type, region, normalised_d2h) %>% 
  mutate(normalised_d2h = as.numeric(normalised_d2h)) %>%
  pivot_longer(
    cols = c(normalised_d2h),
    names_to = "isotope",
    values_to = "value"
  )

CNOH_dat <- 
  bind_rows(O_dat, CN_dat, H_dat)

CNOH_dat %>% 
  group_by(Specimen_ID) %>% 
  summarise(n_isotopes = n_distinct(isotope)) %>% 
  pull(n_isotopes) %>%
  table()
```

### import and wrangle sample meta data

```{r}
# import meta data for all museum samples and add spatial data.
meta_data <-
  read_excel(here("data/MASTER_museum_ALL SAMPLES INVENTORY_banded_dotterels_v4.xlsx"), 
             col_types = "text") %>% 
  mutate(Specimen_ID_ = str_remove(Specimen_ID, regex(" (Breast|Primary)$", ignore_case = TRUE))) %>%
  mutate(Specimen_ID_ = str_remove_all(Specimen_ID_, regex("AJB ", ignore_case = TRUE)))  %>%
  mutate(Specimen_ID_ = str_remove_all(Specimen_ID_, regex(" -", ignore_case = TRUE))) %>% 
  dplyr::select(Feather_type, Specimen_ID, Specimen_ID_, General_collection_location, 
         Specific_collection_location, Collection_date) %>% 
  mutate(Month = str_extract(Collection_date, "(January|February|March|April|May|June|July|August|September|October|November|December)")) %>%
  mutate(Month = match(tolower(Month), tolower(month.name))) %>%
  mutate(
    month_extracted = sapply(Collection_date, extract_month),
    Month = ifelse(is.na(Month) & !is.na(month_extracted), month_extracted, Month),
    Year = sapply(Collection_date, extract_year)
  ) %>%
  dplyr::select(-month_extracted) %>% 
  mutate(location = paste(Specific_collection_location, General_collection_location, sep = ", ")) %>%
  mutate(location = str_replace_all(location, "NA, ", "")) %>% 
  filter(!is.na(Specimen_ID))

meta_data %>% filter(Specimen_ID_ == "AV1740")

# Create a data frame 
df <- data.frame(location = meta_data$location)

# Remove duplicates to avoid redundant API calls
unique_locations <- unique(df$location)

# Get latitude and longitude for unique locations
geocode_data <- geocode(unique_locations)

# Combine the geocode data with the unique locations
location_data <- 
  data.frame(location = unique_locations, geocode_data) %>%
  filter(!is.na(location)) %>% arrange(lat)

meta_data_ <-
  left_join(meta_data, location_data, by = "location") %>% 
  dplyr::select(-c(Feather_type, Specimen_ID)) %>%
  distinct() %>% 
  rename(Specimen_ID = Specimen_ID_)
```

### merge C, N, O, and H data with metadata

```{r}
meta_data_ %>% filter(Specimen_ID == "AV1740")
CNOH_dat %>% filter(Specimen_ID == "AV1740")

CNOH_dat_meta <-
  stringdist_left_join(
    CNOH_dat, meta_data_,
    by = "Specimen_ID",
    method = "cosine", # Jaro-Winkler distance
    # max_dist = 1,         # tweak based on how much difference you allow
    distance_col = "dist"
  ) %>% #filter(Specimen_ID.y == "AV1740" | Specimen_ID.x == "AV1740") %>% group_by(Specimen_ID.x) %>% arrange(dist)
  group_by(Specimen_ID.x, Feather_type, region, isotope) %>% arrange(dist) %>% slice(1) %>%
  dplyr::select(Specimen_ID.x, Specimen_ID.y, dist, everything()) %>% ungroup() %>% arrange(desc(dist)) %>% 
  dplyr::select(-c(Specimen_ID.y, dist)) %>% 
  rename(Specimen_ID = Specimen_ID.x) %>% 
  mutate(geo_region = ifelse(str_detect(location, "Australia"), "Australia", 
                             ifelse(str_detect(location, "New Zealand"), "New Zealand", NA))) %>% 
  mutate(geo_region_bin = ifelse(geo_region == "Australia", 1, 0)) %>%
  mutate(
    month_name = month.name[Month],          # full month names
    month_abbrev = month.abb[Month],         # abbreviated names
    month_factor = factor(Month, levels = 1:12, labels = month.name, ordered = TRUE) # ordered factor
  ) %>% 
  mutate(lon = ifelse(lon < 0, lon + 180*2, lon))
```

### merge with the triplicate data for a full dataset

```{r}
triplicate_CNOH_dat_meta <- 
  triplicate_CNOH_dat_meta %>% 
  mutate(region = "homogenate")

CNOH_dat_meta <- 
  CNOH_dat_meta %>% 
  mutate(feather = NA)

CNOH_dat_meta_full <- 
  bind_rows(triplicate_CNOH_dat_meta, CNOH_dat_meta) %>% 
  dplyr::select(Specimen_ID, Feather_type, feather, region, amount, 
                isotope, lon, lat, geo_region, geo_region_bin, value, Collection_date, Month, Year, Specific_collection_location) %>% 
  group_by(isotope) %>%                # scale within each isotope type
  mutate(value_scaled = as.numeric(scale(value))) %>%
  group_by(Specimen_ID, Feather_type,
                isotope, lon, lat, geo_region, geo_region_bin, Collection_date, Month, Year, Specific_collection_location) %>%
  summarise(mean_value = mean(value_scaled))

saveRDS(CNOH_dat_meta_full, file = "data/CNOH_museum_samples_201125.rds")

# CNOH_dat_meta_full <- 
#   bind_rows(triplicate_CNOH_dat_meta, CNOH_dat_meta) %>% 
#   dplyr::select(Specimen_ID, Feather_type, feather, region, amount, everything()) %>% 
#   group_by(Specimen_ID, Feather_type, isotope, General_collection_location,
#            Specific_collection_location, Month, location, lon, lat, geo_region, 
#            geo_region_bin, month_name, month_abbrev, month_factor) %>% 
#   summarise(mean_value = mean(value))
```

### exploration of CNOH breast feathers

```{r}
CNOH_dat_meta_full <- readRDS(file = "data/CNOH_museum_samples_201125.rds")
```

```{r}
library(dplyr)
library(leaflet)
library(viridisLite)

# Prepare data ----
CNOH_map <- 
  CNOH_dat_meta_full %>% 
  # select(-value_scaled) %>% 
  # group_by(Specimen_ID, isotope, Feather_type) %>% 
  # mutate(mean_value = mean(value)) %>% 
  # dplyr::select(-c(feather, value, amount, value, region)) %>% 
  # distinct() %>% 
  group_by(isotope) %>%                # scale within each isotope type
  mutate(mean_value_scaled = as.numeric(scale(mean_value))) %>%
  # filter(Feather_type == "Breast") %>%
  filter(Month %in% c(5,6,7,8)) %>% 
  # filter((Month %in% c(6,7,) & geo_region == "New Zealand") | (Month %in% c(6,7,8,9,10,11) & geo_region == "Australia")) %>%
  ungroup() %>% 
  group_by(Specimen_ID) %>% 
  mutate(n_feather_types = n_distinct(Feather_type)) %>%
                   mutate(
                     Month = factor(Month, levels = 1:12, labels = month.abb)
                   ) #%>% 
  # filter(n_feather_types == 2)
  # dplyr::select(Specimen_ID, Feather_type, isotope, lon, lat, Month, Year, Specific_collection_location, mean_value) %>% 
  # group_by(isotope) %>%                # scale within each isotope type
  # mutate(mean_value_scaled = as.numeric(scale(mean_value)))

iso_sf <- 
  CNOH_map %>% 
  mutate(isotope = ifelse(isotope == "normalised_d13c", "d13C",
                           ifelse(isotope == "normalised_d15n", "d15N",
                                  ifelse(isotope == "normalised_d2h", "d2H", "d18O")))) %>% 
  filter(Feather_type == "Breast") %>% 
  st_as_sf(., coords = c("lon", "lat"), crs = 4326)
iso_sf <- st_transform(iso_sf, 3857)  # Web Mercator
  
library(ggplot2)
library(sf)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
library(ggforce)

# your sf object: iso_sf
# columns: isotope, mean_value, geometry (POINTS)

# get basemap
world <- ne_countries(scale = "medium", returnclass = "sf")

# restrict map to Australia + New Zealand region
region_bbox <- st_bbox(c(xmin = 110, xmax = 180, ymin = -55, ymax = -0), crs = st_crs(world))
world_region <- st_crop(world, region_bbox)

# example palettes per isotope
isotope_palettes <- list(
  "d13C" = scale_color_viridis_c(option = "plasma", name = expression(delta^13*C[scaled])),
  "d15N" = scale_color_viridis_c(option = "magma", name = expression(delta^15*N[scaled])),
  "d2H"  = scale_color_viridis_c(option = "viridis", name = expression(delta^2*H[scaled])),
  "d18O" = scale_color_viridis_c(option = "cividis", name = expression(delta^18*O[scaled]))
)

# helper to make per-isotope map
make_iso_plot <- function(iso_name, xlim_, y_lim_) {
  data_ = filter(iso_sf, isotope == iso_name) %>%
    mutate(geometry = st_jitter(geometry, amount = 30000))
  
  ggplot() +
    geom_sf(data = world_region, fill = "grey90", color = "white", size = 0.1) +
    geom_sf(
      data = data_,
      aes(color = mean_value_scaled),
      size = 2,
      alpha = 0.8
    ) +
    isotope_palettes[[iso_name]] +
    
    # ---- HORIZONTAL COLORBAR INSIDE PLOT ----
    guides(
      color = guide_colorbar(
        direction = "horizontal",
        title.position = "left",
        barwidth = unit(100, "pt"),
        barheight = unit(10, "pt")
      )
    ) +
    luke_theme +
    theme(
      legend.position = c(0, 0),              # inside panel
      legend.justification = c(0, 0),             # anchor bottom-left
      legend.background = element_blank(),
      legend.key.height = unit(4, "pt"),
      legend.key.width = unit(40, "pt"),
      legend.title = element_text(size = 15),
      
      panel.grid.major = element_line(color = "grey95"),
      axis.text = element_text(size = 8),
      plot.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    
    coord_sf(
      xlim = xlim_,
      ylim = y_lim_,
      expand = FALSE
    )
}

# build and combine plots
AU_NZ_plots <- 
  lapply(unique(iso_sf$isotope), function(x) {
  make_iso_plot(iso_name = x, xlim_ = c(130, 180), y_lim_ = c(-50, -32))
})
final_AU_NZ_plot <- wrap_plots(AU_NZ_plots, ncol = 2)
final_AU_NZ_plot

VIC_TAS_plots <- 
  lapply(unique(iso_sf$isotope), function(x) {
  make_iso_plot(iso_name = x, xlim_ = c(140, 150), y_lim_ = c(-45, -35))
})
final_VIC_TAS_plot <- wrap_plots(VIC_TAS_plots, ncol = 2)
final_VIC_TAS_plot

NZ_plots <- 
  lapply(unique(iso_sf$isotope), function(x) {
  make_iso_plot(iso_name = x, xlim_ = c(165, 180), y_lim_ = c(-47, -35))
})
final_NZ_plot <- wrap_plots(NZ_plots, ncol = 2)
final_NZ_plot


CNOH_map %>% 
      filter(Feather_type == "Breast") %>% 
      filter(!is.na(Month)) %>% 
  # filter(Specific_collection_location == "Ralphs Bay") %>% 
  group_by(isotope, Month) %>% 
    summarise(n_inds = n_distinct(Specimen_ID)) %>% 
  filter(Month == "Jul")

iso_labels <- c(
  "normalised_d13c" = "delta^{13}*C",
  "normalised_d15n" = "delta^{15}*N",
  "normalised_d2h"  = "delta^{2}*H",
  "normalised_d18o" = "delta^{18}*O"
)

ggplot() +
  geom_bar(
    data = CNOH_map %>% 
      filter(Feather_type == "Breast"), #%>% 
      # filter(!is.na(Month)),
    aes(x = Month)
  ) +
  ylab("number of specimens") +
  facet_grid(geo_region ~ isotope,
             labeller = labeller(isotope = as_labeller(iso_labels, label_parsed))) +
  luke_theme  +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines"),
    legend.position = "none",
    
    # Rotate vertical strip text
    strip.text.x = element_text(size = 15),  # 0 = horizontal, 90 = vertical
    strip.background = element_blank()
  )

CNOH_map %>% filter(n_feather_types == 2)

# Initialize leaflet map ----
leaflet_map <- leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles(providers$OpenStreetMap, group = "Terrain") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite")   # background map

# Loop through isotopes ----
for (iso in unique(CNOH_map$isotope)) {
  df <- CNOH_map %>% filter(isotope == iso)
  if (nrow(df) == 0) next  # skip if no data
  
  # Dynamically assign palette
  pal_option <- switch(iso,
                       "normalised_d13c" = "plasma",
                       "normalised_d15n" = "magma",
                       "normalised_d2h"  = "viridis",
                       "normalised_d18o" = "cividis",
                       "viridis") # default fallback
  
  pal <- colorNumeric(
    palette = viridis(100, option = pal_option),
    domain = df$mean_value_scaled
  )
  
  # Add markers and legend for this isotope
  leaflet_map <- leaflet_map %>%
    addCircleMarkers(
      data = df,
      lng = ~lon,
      lat = ~lat,
      color = ~pal(mean_value_scaled),
      radius = 25,
      stroke = FALSE,
      fillOpacity = 0.8,
      popup = ~paste0(
        "<b>Specimen ID:</b> ", Specimen_ID, "<br>",
        "<b>Month:</b> ", Month, "<br>",
        "<b>Year:</b> ", Year, "<br>",
        "<b>Location:</b> ", Specific_collection_location, "<br>",
        "<b>Mean value:</b> ", round(mean_value_scaled, 2), "<br>",
        "<b>Lon / Lat:</b> ", paste0(round(lon, 2), " / ", round(lat, 2)), "<br>",
        "<b>Feather type:</b> ", Feather_type
      ),
      group = iso,
      clusterOptions = markerClusterOptions(
        spiderfyOnMaxZoom = TRUE,
        spiderfyDistanceMultiplier = 3  # <-- longer spiderfy arms
      )
    ) %>%
    addLegend(
      "bottomleft",
      pal = pal,
      values = df$mean_value_scaled,
      title = iso,
      opacity = 1,
      layerId = paste0("legend_", iso)
    )
}

# Add layer control and zoom to bounds ----
leaflet_map <- leaflet_map %>%
  addLayersControl(
    baseGroups = c("Terrain", "Satellite"),
    overlayGroups = unique(CNOH_map$isotope),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  fitBounds(
    lng1 = min(CNOH_map$lon, na.rm = TRUE),
    lat1 = min(CNOH_map$lat, na.rm = TRUE),
    lng2 = max(CNOH_map$lon, na.rm = TRUE),
    lat2 = max(CNOH_map$lat, na.rm = TRUE)
  )

# Display the map ----
leaflet_map
```


```{r}
CNOH_wide <- 
  CNOH_dat_meta_full %>% 
  filter(Feather_type == "Breast") %>%
  ungroup() %>%
  mutate(
    isotope = case_when(
      isotope == "normalised_d13c" ~ "d13C",
      isotope == "normalised_d15n" ~ "d15N",
      isotope == "normalised_d2h"  ~ "d2H",
      isotope == "normalised_d18o" ~ "d18O",
      TRUE ~ isotope
    )
  ) %>%
  # average across repeated isotope measurements per specimen
  # group_by(Specimen_ID, Feather_type, isotope, lon, lat, geo_region_bin, Month) %>%
  # summarise(mean_value = mean(value)) %>%
  group_by(isotope) %>%                # scale within each isotope type
  mutate(mean_value_scaled = as.numeric(scale(mean_value)),
         # feather = ifelse(is.na(feather), "A", feather),
         isotope = as.factor(isotope)) %>%
  ungroup() %>% 
  filter(Month %in% c("5", "6", "7", "8")) %>% 
  select(-mean_value) %>% 
  # pivot to wide format
  pivot_wider(
    names_from = isotope,
    values_from = mean_value_scaled
  )

# sample size
CNOH_dat_meta_full %>%
  filter(Feather_type == "Breast", Month %in% c("5", "6", "7", "8")) %>%
  mutate(isotope_clean = str_remove(isotope, "normalised_")) %>%
  group_by(Specimen_ID) %>%
  summarise(
    combo = paste(sort(unique(isotope_clean)), collapse = "+"),
    n_isotopes = n_distinct(isotope_clean)
  ) %>%
  count(n_isotopes, combo, name = "n_individuals") %>%
  arrange(desc(n_isotopes), desc(n_individuals))

```

#### linear mixed model
```{r}
# CNOH_dat_meta_full_s <- 
#   bind_rows(triplicate_CNOH_dat_meta, CNOH_dat_meta) %>%
#   mutate(
#     isotope = case_when(
#       isotope == "normalised_d13c" ~ "d13C",
#       isotope == "normalised_d15n" ~ "d15N",
#       isotope == "normalised_d2h"  ~ "d2H",
#       isotope == "normalised_d18o" ~ "d18O",
#       TRUE ~ isotope
#     )
#   ) %>% 
#   filter(Feather_type == "Breast", Month %in% c("5", "6", "7", "8")) %>% 
#   select(Specimen_ID, feather, isotope, value, lon, lat, geo_region_bin) %>% 
#   group_by(Specimen_ID, feather, isotope, lon, lat, geo_region_bin) %>% 
#   summarise(mean_value = mean(value)) %>%
#   group_by(isotope) %>%                # scale within each isotope type
#   mutate(mean_value_scaled = as.numeric(scale(mean_value)),
#          feather = ifelse(is.na(feather), "A", feather),
#          isotope = as.factor(isotope)) %>%
#   ungroup()

CNOH_dat_meta_full_s_ <- 
  CNOH_dat_meta_full %>% 
  group_by(isotope) %>%                # scale within each isotope type
  mutate(mean_value_scaled = as.numeric(scale(mean_value))) %>% 
  # group_by(Specimen_ID, isotope, Feather_type) %>% 
  # mutate(mean_value_scaled = mean(value_scaled)) %>% 
  # dplyr::select(-c(feather, value, amount, value_scaled, region)) %>% 
  distinct() %>% 
  filter(Feather_type == "Breast") %>%
  filter((Month %in% c(5,6,7,8) & geo_region == "New Zealand") | (Month %in% c(5,6,7,8,9,10,11,12) & geo_region == "Australia")) %>%
  ungroup() %>% 
  filter(lon > 130)

CNOH_full_mod1 <- 
  glmer(geo_region_bin ~ mean_value_scaled + (1 | isotope), 
       data = CNOH_dat_meta_full_s_, family = binomial)

summary(CNOH_full_mod1)
exp(coef(CNOH_full_mod1))

CNOH_lon_mod2 <- 
  lmer(mean_value_scaled ~ lon + lat + (1 | isotope), 
       data = CNOH_dat_meta_full_s_)

CNOH_reg_lat_full_mod <- 
  lmer(mean_value_scaled ~ geo_region + lat + (1 | isotope),
       data = CNOH_dat_meta_full_s_)
summary(CNOH_reg_lat_full_mod)
plot(allEffects(CNOH_reg_lat_full_mod))

CNOH_reg_lat_CN_mod <- 
  lmer(mean_value_scaled ~ geo_region + lat + (1 | isotope),
       data = CNOH_dat_meta_full_s_ %>% filter(isotope %in% c("normalised_d13c", "normalised_d15n")))
summary(CNOH_reg_lat_CN_mod)
plot(allEffects(CNOH_reg_lat_CN_mod))

CNOH_reg_lat_CN_mod <- 
  glmer(geo_region_bin ~ mean_value_scaled + lat + (1 | isotope),
       data = CNOH_dat_meta_full_s_ %>% filter(isotope %in% c("normalised_d13c", "normalised_d15n")),
       family = binomial)
summary(CNOH_reg_lat_CN_mod)
plot(allEffects(CNOH_reg_lat_CN_mod))

CNOH_reg_lat_CNH_mod <- 
  lmer(mean_value_scaled ~ geo_region + lat + (1 | isotope),
       data = CNOH_dat_meta_full_s_ %>% filter(isotope %in% c("normalised_d13c", "normalised_d15n", "normalised_d2h")))
summary(CNOH_reg_lat_CNH_mod)
plot(allEffects(CNOH_reg_lat_CNH_mod))

# 1️⃣ Create new data for prediction
newdata <- CNOH_dat_meta_full_s_ %>% filter(isotope %in% c("normalised_d13c", "normalised_d15n")) %>%
  group_by(isotope) %>%
  summarise(
    mean_value_scaled = seq(min(mean_value_scaled), max(mean_value_scaled), length.out = 50),
    lat = mean(lat)
  ) %>%
  ungroup()

# 2️⃣ Predict probabilities AND standard errors for CI
pred <- predict(CNOH_reg_lat_CN_mod, newdata, type = "link", se.fit = TRUE)

# Calculate predicted probability and 95% CI
newdata <- newdata %>%
  mutate(
    fit_link = pred$fit,
    se_link = pred$se.fit,
    # logistic transformation
    pred_prob = plogis(fit_link),
    lower = plogis(fit_link - 1.96 * se_link),
    upper = plogis(fit_link + 1.96 * se_link)
  )

# 3️⃣ Plot raw data + predicted probability + confidence intervals
ggplot() +
  # raw data
  geom_jitter(data = CNOH_dat_meta_full_s_ %>% filter(isotope %in% c("normalised_d13c", "normalised_d15n")), 
              aes(x = mean_value_scaled, y = geo_region_bin, color = isotope), 
              width = 0.1, height = 0.02, alpha = 0.6) +
  # predicted probability curves
  geom_line(data = newdata, aes(x = mean_value_scaled, y = pred_prob), size = 1.2) +
  # confidence interval ribbons
  geom_ribbon(data = newdata, aes(x = mean_value_scaled, ymin = lower, ymax = upper), 
              alpha = 0.2, color = NA) +
  theme_minimal() +
  labs(
    x = "Scaled isotope value",
    y = "Probability of having wintered in Australia",
    title = "Predicted probability by scaled value and isotope with raw data"
  )

library(glmmTMB)

# Build a spatial covariance structure using 'exp' (exponential)
glm_fit <- glmmTMB(
  geo_region_bin ~ mean_value_scaled + isotope + 
    (1 | gr(lon + lat, cov = "exp")),  # spatial random effect
  data = CNOH_dat_meta_full_s,
  family = binomial
)
  
install.packages("spaMM")
library(spaMM)

glm_fit <- fitme(
  geo_region_bin ~ mean_value_scaled + isotope + Matern(1 | lon + lat),
  data = CNOH_dat_meta_full_s,
  family = binomial()
)

# For each isotope, create a sequence of mean_value_scaled and include coordinates
# If you have multiple locations per isotope, you can just pick a representative point
# or expand predictions for each existing location

newdata <- CNOH_dat_meta_full_s %>%
  group_by(isotope) %>%
  summarise(
    mean_value_scaled = seq(min(mean_value_scaled), max(mean_value_scaled), length.out = 50),
    lon = mean(lon),   # use mean lon for that isotope
    lat = mean(lat)    # use mean lat for that isotope
  ) %>%
  ungroup()

# Predict
pred <- predict(glm_fit, newdata, type = "response")
newdata$pred_prob <- pred

# Plot
library(ggplot2)

ggplot(newdata, aes(x = mean_value_scaled, y = pred_prob, color = isotope)) +
  geom_line(size = 1.2) +
  theme_minimal() +
  labs(
    x = "Scaled value",
    y = "Predicted probability of geo_region_bin = 1",
    title = "Predicted probabilities by isotope (spaMM model)"
  )
```


```{r}
# install.packages("VIM")
library(VIM)
CNOH_imputed <- kNN(CNOH_wide, variable = c("d13C","d15N","d2H","d18O"), k = 5)
```

#### specimens with all 4 isotopes
```{r}
# run PCA on isotopes
CNOH_wide_no_nas <- CNOH_wide %>% na.omit()

pca <- prcomp(CNOH_wide_no_nas[, c("d13C", "d15N", "d2H", "d18O")], scale. = TRUE)

# add PC scores
iso_pca <- cbind(CNOH_wide_no_nas, pca$x[, 1:2])

# logistic model on PC axes
m_pca <- glm(geo_region_bin ~ PC1 + PC2, data = iso_pca, family = binomial)
summary(m_pca)

ggplot(iso_pca, aes(PC1, PC2, color = as.factor(geo_region_bin))) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(color = "Country (1=AU)")


library(MASS)

# Remove rows with any missing isotopes
iso_lda <- CNOH_wide %>%
  drop_na(d13C, d15N, d2H, d18O)

# Make geo_region_bin a factor (required by lda)
iso_lda <- iso_lda %>%
  mutate(region = factor(geo_region_bin, levels = c(0, 1), labels = c("NZ", "AU")))

# Run LDA
lda_model <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda)

# View results
lda_model

# Get predictions
pred <- predict(lda_model)

# Add to dataframe
iso_lda <- iso_lda %>%
  mutate(LD1 = pred$x[, 1],
         predicted = pred$class)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = iso_lda$predicted)

# Classification accuracy
mean(iso_lda$region == iso_lda$predicted)

ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "LDA separation between AU and NZ individuals",
       x = "Linear Discriminant 1", fill = "Region")

ggplot(iso_lda, aes(x = LD1, y = 0, color = region)) +
  geom_jitter(width = 0, height = 0.05, size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(x = "LD1", y = NULL, title = "LDA projection")

lda_cv <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda, CV = TRUE)

# Accuracy
mean(lda_cv$class == iso_lda$region)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = lda_cv$class)

ggplot(iso_lda, aes(x = region, y = LD1, color = region)) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", size = 4, shape = 23, fill = "white") +
  scale_color_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  labs(title = "LDA Separation by Region",
       y = "LD1 score",
       x = "Region")

# Assuming lda_model is already fitted
lda_coeff <- lda_model$scaling[, 1]  # LD1 coefficients
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_coeff
) %>%
  arrange(desc(abs(coefficient)))  # order by absolute contribution

ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = "none") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- LDA model (already fitted) ---
lda_model <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda)

# --- LD1 scores ---
lda_pred <- predict(lda_model)
lda_df <- data.frame(CNOH_wide, LD1 = lda_pred$x[, 1])

# --- LD1 density plot ---
p1 <- ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "LDA separation by region",
    x = "LD1",
    y = "Density",
    fill = "Region"
  )

# --- Extract LD1 coefficients (loadings) ---
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_model$scaling[, 1]
) %>%
  arrange(desc(abs(coefficient)))

# --- Isotope contribution bar plot ---
p2 <- ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- Combine plots with patchwork ---
combined_plot <- p1 / p2 + plot_layout(heights = c(2, 1))
combined_plot

library(caret)

set.seed(123)
train_control <- trainControl(method = "LOOCV")

lda_cv <- train(
  as.factor(geo_region_bin) ~ d13C + d15N + d2H + d18O,
  data = CNOH_wide_no_nas,
  method = "lda",
  trControl = train_control
)

lda_cv
```

#### specimens with at least 2 isotopes, imputed based on spatial proximity
```{r}
# install.packages("mice")
library(mice)
CNOH_wide <- 
  CNOH_dat_meta_full_s_ %>%
  select(-mean_value) %>% 
  pivot_wider(
    names_from  = isotope,
    values_from = mean_value_scaled
  ) %>% 
  rename(d13C = normalised_d13c, 
         d15N = normalised_d15n, 
         d2H = normalised_d2h, 
         d18O = normalised_d18o) %>% 
  distinct()

mice_data <- CNOH_wide %>%
  dplyr::select(d13C, d15N, d2H, d18O, geo_region_bin, lon, lat)

mice_imp <- mice(mice_data, m = 5, method = "pmm", seed = 123)  # Predictive mean matching
CNOH_imputed <- complete(mice_imp, 1)  # use first imputed dataset

# run PCA on isotopes
pca <- prcomp(CNOH_imputed[, c("d13C", "d15N", "d2H", "d18O")], scale. = TRUE)

# add PC scores
iso_pca <- cbind(CNOH_imputed, pca$x[, 1:2])

# logistic model on PC axes
m_pca <- glm(geo_region_bin ~ PC1 + PC2, data = iso_pca, family = binomial)
summary(m_pca)

ggplot(iso_pca, aes(PC1, PC2, color = as.factor(geo_region_bin))) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(color = "Country (1=AU)")

library(MASS)

# Make geo_region_bin a factor (required by lda)
iso_lda <- CNOH_imputed %>%
  mutate(region = factor(geo_region_bin, levels = c(0, 1), labels = c("NZ", "AU")))

# Run LDA
lda_model <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda)

# View results
lda_model

# Get predictions
pred <- predict(lda_model)

# Add to dataframe
iso_lda <- iso_lda %>%
  mutate(LD1 = pred$x[, 1],
         predicted = pred$class)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = iso_lda$predicted)

# Classification accuracy
mean(iso_lda$region == iso_lda$predicted)

ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "LDA separation between AU and NZ individuals",
       x = "Linear Discriminant 1", fill = "Region")

ggplot(iso_lda, aes(x = LD1, y = 0, color = region)) +
  geom_jitter(width = 0, height = 0.05, size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(x = "LD1", y = NULL, title = "LDA projection")

lda_cv <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda, CV = TRUE)

# Accuracy
mean(lda_cv$class == iso_lda$region)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = lda_cv$class)

ggplot(iso_lda %>% mutate(region = factor(region, levels = c("AU", "NZ"))), aes(x = region, y = LD1, color = region)) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", size = 4, shape = 23, fill = "white") +
  scale_color_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("NZ" = "New Zealand",
                              "AU" = "Australia")) +
  labs(y = "Linear discriminant axis 1\n scaled isotopic signature of δ13C, δ15N, δ18O, δ2H",
       x = "collection location in winter")

# Assuming lda_model is already fitted
lda_coeff <- lda_model$scaling[, 1]  # LD1 coefficients
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_coeff
) %>%
  arrange(desc(abs(coefficient)))  # order by absolute contribution

ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = "none") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- LDA model (already fitted) ---
lda_model <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda)

# --- LD1 scores ---
lda_pred <- predict(lda_model)
lda_df <- data.frame(CNOH_wide, LD1 = lda_pred$x[, 1])

# --- LD1 density plot ---
p1 <- ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "LDA separation by region",
    x = "LD1",
    y = "Density",
    fill = "Region"
  )

# --- Extract LD1 coefficients (loadings) ---
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_model$scaling[, 1]
) %>%
  arrange(desc(abs(coefficient)))

# --- Isotope contribution bar plot ---
p2 <- ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- Combine plots with patchwork ---
combined_plot <- p1 / p2 + plot_layout(heights = c(2, 1))
combined_plot

library(caret)

set.seed(123)
train_control <- trainControl(method = "LOOCV")

lda_cv <- train(
  as.factor(geo_region_bin) ~ d13C + d15N + d2H + d18O,
  data = CNOH_imputed,
  method = "lda",
  trControl = train_control
)

lda_cv

```

#### only C and N
#### specimens with at least 2 isotopes, imputed based on spatial proximity
```{r}
# install.packages("mice")
library(mice)

mice_data <- CNOH_wide %>%
  dplyr::select(d13C, d15N, geo_region_bin, lon, lat)

mice_imp <- mice(mice_data, m = 5, method = "pmm", seed = 123)  # Predictive mean matching
CNOH_imputed <- complete(mice_imp, 1)  # use first imputed dataset

# run PCA on isotopes
pca <- prcomp(CNOH_imputed[, c("d13C", "d15N")], scale. = TRUE)

# add PC scores
iso_pca <- cbind(CNOH_imputed, pca$x[, 1:2])

# logistic model on PC axes
m_pca <- glm(geo_region_bin ~ PC1 + PC2, data = iso_pca, family = binomial)
summary(m_pca)

ggplot(iso_pca, aes(PC1, PC2, color = as.factor(geo_region_bin))) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(color = "Country (1=AU)")


library(MASS)

# Make geo_region_bin a factor (required by lda)
iso_lda <- CNOH_imputed %>%
  mutate(region = factor(geo_region_bin, levels = c(0, 1), labels = c("NZ", "AU")))

# Run LDA
lda_model <- lda(region ~ d13C + d15N, data = iso_lda)

# View results
lda_model

# Get predictions
pred <- predict(lda_model)

# Add to dataframe
iso_lda <- iso_lda %>%
  mutate(LD1 = pred$x[, 1],
         predicted = pred$class)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = iso_lda$predicted)

# Classification accuracy
mean(iso_lda$region == iso_lda$predicted)

ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "LDA separation between AU and NZ individuals",
       x = "Linear Discriminant 1", fill = "Region")

ggplot(iso_lda, aes(x = LD1, y = 0, color = region)) +
  geom_jitter(width = 0, height = 0.05, size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(x = "LD1", y = NULL, title = "LDA projection")

lda_cv <- lda(region ~ d13C + d15N, data = iso_lda, CV = TRUE)

# Accuracy
mean(lda_cv$class == iso_lda$region)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = lda_cv$class)

ggplot(iso_lda %>% mutate(region = factor(region, levels = c("AU", "NZ"))), aes(x = region, y = LD1, color = region)) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", size = 4, shape = 23, fill = "white") +
  scale_color_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("NZ" = "New Zealand",
                              "AU" = "Australia")) +
  labs(y = "Linear discriminant axis 1\nIsotopic signature of δ13C, δ15N, δ18O, δ2H",
       x = "collection location in winter")

# Assuming lda_model is already fitted
lda_coeff <- lda_model$scaling[, 1]  # LD1 coefficients
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_coeff
) %>%
  arrange(desc(abs(coefficient)))  # order by absolute contribution

ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = "none") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- LDA model (already fitted) ---
lda_model <- lda(region ~ d13C + d15N, data = iso_lda)

# --- LD1 scores ---
lda_pred <- predict(lda_model)
lda_df <- data.frame(CNOH_wide, LD1 = lda_pred$x[, 1])

# --- LD1 density plot ---
p1 <- ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "LDA separation by region",
    x = "LD1",
    y = "Density",
    fill = "Region"
  )

# --- Extract LD1 coefficients (loadings) ---
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_model$scaling[, 1]
) %>%
  arrange(desc(abs(coefficient)))

# --- Isotope contribution bar plot ---
p2 <- ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- Combine plots with patchwork ---
combined_plot <- p1 / p2 + plot_layout(heights = c(2, 1))
combined_plot

library(caret)

set.seed(123)
train_control <- trainControl(method = "LOOCV")

lda_cv <- train(
  as.factor(geo_region_bin) ~ d13C + d15N + d2H + d18O,
  data = CNOH_imputed,
  method = "lda",
  trControl = train_control
)

lda_cv

```




```{r}
library(spdep)

coords <- CNOH_wide %>%
  mutate(
    lon_jit = lon + runif(n(), -1e-5, 1e-5),
    lat_jit = lat + runif(n(), -1e-5, 1e-5)
  ) %>%
  dplyr::select(lon_jit, lat_jit)
nb <- knn2nb(knearneigh(coords, k = 1))  # 3 nearest neighbors
spatial_mean <- sapply(1:nrow(coords), function(i) {
  mean(CNOH_wide$d13C[nb[[i]]], na.rm = TRUE)
})

CNOH_mice$spatial_d13C <- spatial_mean
```

### Linear Discrimanent Analysis


### build isoscapes

currently not enough data for this to work

```{r, eval=FALSE}
library(sf)         # spatial objects
library(gstat)      # geostatistics (kriging, variograms)
library(raster)     # raster operations
library(terra)      # modern replacement for raster
library(spData)     # world basemap (optional)

iso_sf <- 
  CNOH_dat_meta_full %>% 
  mutate(isotope = ifelse(isotope == "normalised_d13c", "d13C",
                           ifelse(isotope == "normalised_d15n", "d15N",
                                  ifelse(isotope == "normalised_d2h", "d2H", "d18O")))) %>% 
  filter(Feather_type == "Breast") %>% filter(Month %in% c("5", "6", "7", "8")) %>% 
  group_by(Specimen_ID, geo_region_bin, isotope, lon, lat) %>%
  summarise(mean_value = mean(mean_value, na.rm = TRUE), .groups = "drop") %>% 
  st_as_sf(., coords = c("lon", "lat"), crs = 4326)
iso_sf <- st_transform(iso_sf, 3857)  # Web Mercator
# iso_sf <- 
#   iso_sf %>% 
#   filter(Feather_type == "Breast") %>% filter(Month %in% c("5", "6", "7", "8"))
  

library(ggplot2)
library(sf)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
library(ggforce)

# your sf object: iso_sf
# columns: isotope, mean_value, geometry (POINTS)

# get basemap
world <- ne_countries(scale = "medium", returnclass = "sf")

# restrict map to Australia + New Zealand region
region_bbox <- st_bbox(c(xmin = 110, xmax = 180, ymin = -55, ymax = -0), crs = st_crs(world))
world_region <- st_crop(world, region_bbox)

# example palettes per isotope
isotope_palettes <- list(
  "d13C" = scale_color_viridis_c(option = "plasma", name = expression(delta^13*C)),
  "d15N" = scale_color_viridis_c(option = "magma", name = expression(delta^15*N)),
  "d2H"  = scale_color_viridis_c(option = "viridis", name = expression(delta^2*H)),
  "d18O" = scale_color_viridis_c(option = "cividis", name = expression(delta^18*O))
)

# spread_points <- function(sf_obj, amount = 0.05) {
#   coords <- st_coordinates(sf_obj)
#   dup_groups <- as.data.frame(coords) %>% 
#     dplyr::mutate(id = seq_len(nrow(.))) %>% 
#     dplyr::group_by(X, Y) %>% 
#     dplyr::mutate(
#       n = n(),
#       angle = seq(0, 2*pi, length.out = n + 1)[1:n],
#       X = X + amount * cos(angle),
#       Y = Y + amount * sin(angle)
#     )
#   sf_obj$geometry <- st_as_sf(dup_groups, coords = c("X", "Y"), crs = st_crs(sf_obj))$geometry
#   sf_obj
# }
# 
# make_iso_plot <- function(iso_name) {
#   iso_data <- iso_sf %>%
#     filter(isotope == iso_name) %>%
#     spread_points(amount = 10)
# 
#   ggplot() +
#     geom_sf(data = world_region, fill = "grey90", color = "white", size = 0.1) +
#     geom_sf(data = iso_data, aes(color = mean_value), size = 2, alpha = 0.8) +
#     isotope_palettes[[iso_name]] +
#     coord_sf(xlim = c(110, 180), ylim = c(-55, -20), expand = FALSE) +
#     theme_minimal() +
#     theme(
#       panel.grid.major = element_line(color = "grey95"),
#       axis.text = element_text(size = 8),
#       plot.title = element_text(face = "bold")
#     ) +
#     labs(title = iso_name, x = NULL, y = NULL)
# }

# helper to make per-isotope map
make_iso_plot <- function(iso_name, xlim_, y_lim_) {
  data_ = filter(iso_sf, isotope == iso_name) %>%
    mutate(geometry = st_jitter(geometry, amount = 30000))
  
  ggplot() +
    geom_sf(data = world_region, fill = "grey90", color = "white", size = 0.1) +
    geom_sf(
      data = data_,
      aes(color = mean_value),
      size = 2,
      alpha = 0.8
    ) +
    isotope_palettes[[iso_name]] +
    
    # ---- HORIZONTAL COLORBAR INSIDE PLOT ----
    guides(
      color = guide_colorbar(
        direction = "horizontal",
        title.position = "top",
        barwidth = unit(150, "pt"),
        barheight = unit(10, "pt")
      )
    ) +
    luke_theme +
    theme(
      legend.position = c(0.05, 0.05),              # inside panel
      legend.justification = c(0, 0),             # anchor bottom-left
      legend.background = element_blank(),
      legend.key.height = unit(4, "pt"),
      legend.key.width = unit(40, "pt"),
      legend.title = element_text(size = 15),
      
      panel.grid.major = element_line(color = "grey95"),
      axis.text = element_text(size = 8),
      plot.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    
    coord_sf(
      xlim = xlim_,
      ylim = y_lim_,
      expand = FALSE
    )
}

# build and combine plots
AU_NZ_plots <- 
  lapply(unique(iso_sf$isotope), function(x) {
  make_iso_plot(iso_name = x, xlim_ = c(135, 180), y_lim_ = c(-50, -35))
})
final_AU_NZ_plot <- wrap_plots(AU_NZ_plots, ncol = 2)
final_AU_NZ_plot

VIC_TAS_plots <- 
  lapply(unique(iso_sf$isotope), function(x) {
  make_iso_plot(iso_name = x, xlim_ = c(140, 150), y_lim_ = c(-45, -35))
})
final_VIC_TAS_plot <- wrap_plots(VIC_TAS_plots, ncol = 2)
final_VIC_TAS_plot

NZ_plots <- 
  lapply(unique(iso_sf$isotope), function(x) {
  make_iso_plot(iso_name = x, xlim_ = c(165, 180), y_lim_ = c(-47, -35))
})
final_NZ_plot <- wrap_plots(NZ_plots, ncol = 2)
final_NZ_plot

# bounding box expanded by ~1 degree
bb <- st_bbox(iso_sf_)
grid <- st_make_grid(
  iso_sf_,
  cellsize = 20000, # grid resolution in meters (~20 km)
  what = "centers"
) %>%
  st_as_sf()

# Convert to sp for gstat
iso_sp <- as(iso_sf_, "Spatial")

# ---------------- δ13C ----------------
# Variogram
vgm_d13c <- variogram(value ~ 1, iso_sp[iso_sp@data$isotope == "normalised_d13c", ])
fit_vgm_d13c <- fit.variogram(vgm_d13c, model = vgm("Sph"))

# Kriging
gstat_d13c <- gstat(id = "d13c", formula = value ~ 1, 
                    data = iso_sp[iso_sp@data$isotope == "normalised_d13c", ], 
                    model = fit_vgm_d13c)

grid_sp <- as(grid, "Spatial")
kriged_d13c <- predict(gstat_d13c, newdata = grid_sp)

# Convert to raster
r_d13c <- rasterFromXYZ(as.data.frame(kriged_d13c)[, c("coords.x1", "coords.x2", "d13c.pred")])
plot(r_d13c, main = "Isoscape: δ13C")
points(iso_sp, pch = 20, col = "black")

# ---------------- δ15N ----------------
# Variogram
vgm_d15n <- variogram(value ~ 1, iso_sp[iso_sp@data$isotope == "normalised_d15n", ])
fit_vgm_d15n <- fit.variogram(vgm_d15n, model = vgm("Sph"))

# Kriging
gstat_d15n <- gstat(id = "d15n", formula = value ~ 1, 
                    data = iso_sp[iso_sp@data$isotope == "normalised_d15n", ], 
                    model = fit_vgm_d15n)

kriged_d15n <- predict(gstat_d15n, newdata = grid_sp)

# Raster
r_d15n <- rasterFromXYZ(as.data.frame(kriged_d15n)[, c("coords.x1", "coords.x2", "d15n.pred")])
plot(r_d15n, main = "Isoscape: δ15N")
points(iso_sp[iso_sp@data$isotope == "normalised_d15n", ], pch = 20, col = "black")

# ---------------- δ18O ----------------
# Variogram
vgm_d18o <- variogram(value ~ 1, iso_sp[iso_sp@data$isotope == "normalised_d18o", ])
fit_vgm_d18o <- fit.variogram(vgm_d18o, model = vgm("Gau"))

# Kriging
gstat_d18o <- gstat(id = "d18o", formula = value ~ 1, 
                    data = iso_sp[iso_sp@data$isotope == "normalised_d18o", ], 
                    model = fit_vgm_d18o)

kriged_d18o <- predict(gstat_d18o, newdata = grid_sp)

# Raster
r_d18o <- rasterFromXYZ(as.data.frame(kriged_d18o)[, c("coords.x1", "coords.x2", "d18o.pred")])
plot(r_d18o, main = "Isoscape: δ18O")
points(iso_sp[iso_sp@data$isotope == "normalised_d18o", ], pch = 20, col = "black")
```

## linear models of spatial variation

```{r}
CNOH_dat_meta_breast_winter <- 
  CNOH_dat_meta_full %>% 
  filter(Feather_type == "Breast") %>% 
  filter(Month %in% c("5", "6", "7", "8"))

# δ13C
mod_d13c <- glm(geo_region_bin ~ mean_value,
                data = subset(CNOH_dat_meta_breast_winter, isotope == "normalised_d13c"),
                family = binomial)

summary(mod_d13c)
anova(mod_d13c, test = "Chisq")

# δ15N
mod_d15n <- glm(geo_region_bin ~ mean_value,
                data = subset(CNOH_dat_meta_breast_winter, isotope == "normalised_d15n"),
                family = binomial)

summary(mod_d15n)
anova(mod_d15n, test = "Chisq")

# δ18O
mod_d18o <- glm(geo_region_bin ~ mean_value,
                data = subset(CNOH_dat_meta_breast_winter, isotope == "normalised_d18o"),
                family = binomial)

summary(mod_d18o)
anova(mod_d18o, test = "Chisq")

# δ2H
mod_d2h <- glm(geo_region_bin ~ mean_value,
                data = subset(CNOH_dat_meta_breast_winter, isotope == "normalised_d2h"),
                family = binomial)

summary(mod_d2h)
anova(mod_d2h, test = "Chisq")
```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# --- Prepare single-isotope predicted data ---
predict_iso <- function(mod, data, iso_name){
  pred <- predict(mod, type = "link", se.fit = TRUE)
  data %>%
    ungroup() %>%
    mutate(
      linear_pred = pred$fit,
      fit = plogis(pred$fit),
      lower = plogis(pred$fit - 1.96*pred$se.fit),
      upper = plogis(pred$fit + 1.96*pred$se.fit),
      isotope = iso_name
    )
}

# δ13C
d13c_data <- CNOH_dat_meta_breast_winter %>%
  filter(isotope == "normalised_d13c") %>%
  mutate(geo_region_bin = factor(geo_region_bin, levels = c(0,1)))

mod_d13c <- glm(geo_region_bin ~ mean_value, data = d13c_data, family = binomial)
pred_d13c <- predict_iso(mod_d13c, d13c_data, "δ13C")

# δ15N
d15n_data <- CNOH_dat_meta_breast_winter %>%
  filter(isotope == "normalised_d15n") %>%
  mutate(geo_region_bin = factor(geo_region_bin, levels = c(0,1)))

mod_d15n <- glm(geo_region_bin ~ mean_value, data = d15n_data, family = binomial)
pred_d15n <- predict_iso(mod_d15n, d15n_data, "δ15N")

# δ18O
d18o_data <- CNOH_dat_meta_breast_winter %>%
  filter(isotope == "normalised_d18o") %>%
  mutate(geo_region_bin = factor(geo_region_bin, levels = c(0,1)))

mod_d18o <- glm(geo_region_bin ~ mean_value, data = d18o_data, family = binomial)
pred_d18o <- predict_iso(mod_d18o, d18o_data, "δ18O")

# δ2H
d2h_data <- CNOH_dat_meta_breast_winter %>%
  filter(isotope == "normalised_d2h") %>%
  mutate(geo_region_bin = factor(geo_region_bin, levels = c(0,1)))

mod_d2h <- glm(geo_region_bin ~ mean_value, data = d2h_data, family = binomial)
pred_d2h <- predict_iso(mod_d2h, d2h_data, "δ2H")

# --- Prepare multi-isotope model ---
iso_wide <- CNOH_dat_meta_breast_winter %>%
  pivot_wider(id_cols = c("Specimen_ID","geo_region_bin"),
              names_from = "isotope",
              values_from = "mean_value") %>%
  filter(!is.na(normalised_d13c) & !is.na(normalised_d15n) & !is.na(normalised_d18o) & !is.na(normalised_d2h)) %>%
  mutate(geo_region_bin = factor(geo_region_bin, levels = c(0,1))) %>%
  ungroup()

mod_multi <- glm(geo_region_bin ~ normalised_d13c + normalised_d15n + normalised_d18o + normalised_d2h,
                 data = iso_wide, family = binomial)

pred_multi <- predict(mod_multi, type = "link", se.fit = TRUE)

iso_wide_pred <- iso_wide %>%
  mutate(
    linear_pred = pred_multi$fit,
    fit = plogis(pred_multi$fit),
    lower = plogis(pred_multi$fit - 1.96*pred_multi$se.fit),
    upper = plogis(pred_multi$fit + 1.96*pred_multi$se.fit),
    isotope = "multi-isotope"
  )

# --- Combine all predicted data ---
pred_all <- bind_rows(
  pred_d13c %>% dplyr::select(mean_value, fit, lower, upper, isotope),
  pred_d15n %>% dplyr::select(mean_value, fit, lower, upper, isotope),
  pred_d18o %>% dplyr::select(mean_value, fit, lower, upper, isotope),
  pred_d2h %>% dplyr::select(mean_value, fit, lower, upper, isotope)#,
  # iso_wide_pred %>% dplyr::select(linear_pred, fit, lower, upper, isotope) %>%
    # rename(mean_value = linear_pred)
)

# Combine raw data points for plotting
raw_all <- bind_rows(
  d13c_data %>% dplyr::select(mean_value, geo_region_bin) %>% mutate(isotope = "δ13C"),
  d15n_data %>% dplyr::select(mean_value, geo_region_bin) %>% mutate(isotope = "δ15N"),
  d18o_data %>% dplyr::select(mean_value, geo_region_bin) %>% mutate(isotope = "δ18O"),
  d2h_data %>% dplyr::select(mean_value, geo_region_bin) %>% mutate(isotope = "δ2H")#,
  # iso_wide_pred %>% dplyr::select(linear_pred, geo_region_bin) %>% mutate(isotope = "multi-isotope") %>%
    # rename(mean_value = linear_pred)
)

# Thresholds for plotting
thresholds <- data.frame(
  isotope = c("δ13C","δ15N","δ18O","δ2H"), #,"multi-isotope"),
  threshold = c(0.475, 0.366, 0.282, 0.59)
)

# --- Plot ---
ggplot() +
  geom_point(data = raw_all, aes(x = mean_value, y = as.numeric(geo_region_bin)-1),
             position = position_jitter(height = 0.05), alpha = 0.6) +
  geom_line(data = pred_all, aes(x = mean_value, y = fit), color = "steelblue", size = 1) +
  geom_ribbon(data = pred_all, aes(x = mean_value, ymin = lower, ymax = upper),
              fill = "steelblue", alpha = 0.2) +
  geom_hline(data = thresholds, aes(yintercept = threshold), linetype = "dashed", color = "red") +
  facet_wrap(~isotope, scales = "free_x") +
  labs(x = "Stable isotope value",
       y = "Probability wintering in Australia",
       title = "Predicted effects of single-isotope models on wintering region") +
  theme_minimal()
```

```{r}
CNOH_dat_meta_breast_winter_wide <- tidyr::pivot_wider(
  CNOH_dat_meta_breast_winter,
  id_cols = c(Specimen_ID, geo_region_bin),
  names_from = isotope,
  values_from = mean_value
)

mod_multi <- glm(geo_region_bin ~ normalised_d13c + normalised_d15n + normalised_d18o,
                 data = CNOH_dat_meta_breast_winter_wide,
                 family = binomial)

summary(mod_multi)
anova(mod_multi, test = "Chisq")
```

```{r}
library(pscl)
pR2(mod_d13c)   # univariate R2
pR2(mod_d15n)
pR2(mod_d18o)
pR2(mod_multi)  # multivariate R2

```

## bootstrap ROC analysis
```{r}
install.packages("caret")
install.packages("pROC")
library(caret)
library(pROC)
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r, eval=FALSE}
set.seed(123)
n_runs <- 1000
isotopes <- c("normalised_d13c", "normalised_d15n", "normalised_d18o")

all_iso_results <- list()
threshold_results <- data.frame()

ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

for(run in 1:n_runs){
  
  # --- Single-isotope models ---
  for(iso in isotopes){
    iso_data <- subset(CNOH_dat_meta_breast_winter, isotope == iso)
    iso_data$geo_region_bin <- factor(iso_data$geo_region_bin, levels = c(0,1), labels = c("NZ","AU"))
    
    cv_mod <- train(
      geo_region_bin ~ mean_value,
      data = iso_data,
      method = "glm",
      family = binomial,
      trControl = ctrl,
      metric = "ROC"
    )
    
    pred <- cv_mod$pred$NZ
    roc_obj <- roc(cv_mod$pred$obs, pred, levels = c("AU","NZ"))
    
    roc_df <- coords(roc_obj, "all", ret = c("specificity","sensitivity"))
    roc_df <- as.data.frame(roc_df) %>% mutate(model = iso, run = run)
    all_iso_results[[length(all_iso_results)+1]] <- roc_df
    
    opt <- coords(roc_obj, "best", ret = c("threshold","sensitivity","specificity"), best.method = "youden")
    threshold_results <- rbind(threshold_results,
                               data.frame(run = run, model = iso,
                                          threshold = opt["threshold"],
                                          sensitivity = opt["sensitivity"],
                                          specificity = opt["specificity"]))
  }
  
  # --- Multi-isotope model ---
  iso_wide <- tidyr::pivot_wider(
    CNOH_dat_meta_breast_winter,
    id_cols = c("Specimen_ID","geo_region_bin"),
    names_from = "isotope",
    values_from = "mean_value"
  ) %>%
    filter(!is.na(normalised_d13c) & !is.na(normalised_d15n) & !is.na(normalised_d18o))
  
  iso_wide$geo_region_bin <- factor(iso_wide$geo_region_bin, levels = c(0,1), labels = c("NZ","AU"))
  
  cv_multi <- train(
    geo_region_bin ~ normalised_d13c + normalised_d15n + normalised_d18o,
    data = iso_wide,
    method = "glm",
    family = binomial,
    trControl = ctrl,
    metric = "ROC"
  )
  
  pred_multi <- cv_multi$pred$NZ
  roc_multi <- roc(cv_multi$pred$obs, pred_multi, levels = c("AU","NZ"))
  
  roc_df_multi <- coords(roc_multi, "all", ret = c("specificity","sensitivity"))
  roc_df_multi <- as.data.frame(roc_df_multi) %>% mutate(model = "multi_isotope", run = run)
  all_iso_results[[length(all_iso_results)+1]] <- roc_df_multi
  
  opt_multi <- coords(roc_multi, "best", ret = c("threshold","sensitivity","specificity"), best.method = "youden")
  threshold_results <- rbind(threshold_results,
                             data.frame(run = run, model = "multi_isotope",
                                        threshold = opt_multi["threshold"],
                                        sensitivity = opt_multi["sensitivity"],
                                        specificity = opt_multi["specificity"]))
}

# Combine ROC data
roc_all <- list(iso_results = bind_rows(all_iso_results),
                threshold_results = threshold_results)

saveRDS(roc_all, "output/roc_all.rds")
```


```{r}
roc_all <- readRDS("output/roc_all.rds")

# Compute mean ROC for each model
roc_mean <- roc_all$iso_results %>%
  group_by(model, specificity) %>%
  summarise(sensitivity_mean = mean(sensitivity), .groups = "drop")

# Summarize optimal thresholds with mean and 95% CI
threshold_summary <- roc_all$threshold_results %>%
  group_by(model) %>%
  summarise(
    threshold_mean = mean(threshold),
    threshold_lower = quantile(threshold, 0.025),
    threshold_upper = quantile(threshold, 0.975),
    sensitivity_mean = mean(sensitivity),
    sensitivity_lower = quantile(sensitivity, 0.025),
    sensitivity_upper = quantile(sensitivity, 0.975),
    specificity_mean = mean(specificity),
    specificity_lower = quantile(specificity, 0.025),
    specificity_upper = quantile(specificity, 0.975)
  )

# Convert threshold values to a nicely formatted string
threshold_summary <- threshold_summary %>%
  mutate(threshold_label = sprintf("%.2f", threshold_mean))

threshold_summary

# Plot ROC curves with faint lines + mean curve + diagonal + mean optimal threshold points + labels + 95% CI
ggplot() +
  geom_line(data = roc_all$iso_results %>% group_by(model, run, sensitivity), aes(x = 1 - specificity, y = sensitivity, group = run),
            alpha = 0.1, color = "lightsteelblue") +
  geom_line(data = roc_mean, aes(x = 1 - specificity, y = sensitivity_mean),
            color = "steelblue", size = 2) +
  geom_point(data = threshold_summary, 
             aes(x = 1 - specificity_mean, y = sensitivity_mean),
             color = "black", size = 3) +
  # geom_errorbar(data = threshold_summary,
  #               aes(x = 1 - specificity_mean, ymin = sensitivity_lower, ymax = sensitivity_upper),
  #               color = "black", width = 0.01) +
  # geom_errorbarh(data = threshold_summary,
  #                aes(y = sensitivity_mean, xmin = 1 - specificity_upper, xmax = 1 - specificity_lower),
  #                color = "black", height = 0.01) +
  geom_text(data = threshold_summary,
            aes(x = 1 - specificity_mean, y = sensitivity_mean, label = threshold_label),
            vjust = -1, hjust = 1.1, color = "black", size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  facet_wrap(~model) +
  labs(x = "False Positive Rate (1 - Specificity)", 
       y = "True Positive Rate (Sensitivity)",
       title = "Bootstrap CV ROC curves (1000 runs) with Mean ROC, Optimal Thresholds, and 95% CI") +
  theme_minimal()
```

## Overview

The table summarizes the predictive performance of logistic regression models classifying samples into `geo_region` (NZ vs AU) based on stable isotope values (δ¹³C, δ¹⁵N, δ¹⁸O) and a combined multi-isotope model. For each model, the **mean optimal threshold** (based on Youden’s index) is reported, along with **95% confidence intervals** for threshold, sensitivity, and specificity across 1000 bootstrap cross-validation runs.

## Key Observations

### Multi-isotope model

-   **Mean optimal threshold:** 0.59 (95% CI: 0.29–0.999)\
-   **Sensitivity (true positive rate for AU):** 0.76 (CI: 0.50–0.83)\
-   **Specificity (true negative rate for NZ):** 0.73 (CI: 0.50–1)

**Interpretation:** Combining δ¹³C, δ¹⁵N, and δ¹⁸O provides moderately balanced performance, with some uncertainty due to small sample size, but achieves reasonably high sensitivity and specificity simultaneously.

### δ¹³C (`normalised_d13c`)

-   **Mean threshold:** 0.475 (CI: 0.418–0.554)\
-   **Sensitivity (AU correctly identified):** 0.78 (CI: 0.778–0.778)\
-   **Specificity (NZ correctly identified):** 0.915 (CI: 0.824–0.941)

**Interpretation:** δ¹³C alone is highly specific—good at correctly identifying NZ samples—but slightly less sensitive in detecting AU samples. Narrow CI suggests consistent performance across bootstrap runs.\
- **Context:** ROC = 0.93 → excellent discriminative ability of δ¹³C for separating NZ vs AU samples. Sensitivity = 0.78 → the model correctly identifies 78% of AU samples. Specificity = 0.915 → the model correctly identifies 91.5% of NZ samples.

### δ¹⁵N (`normalised_d15n`)

-   **Mean threshold:** 0.366 (CI: 0.256–0.441)\
-   **Sensitivity (AU correctly identified):** 0.68 (CI: 0.556–0.889)\
-   **Specificity (NZ correctly identified):** 0.78 (CI: 0.529–0.882)

**Interpretation:** δ¹⁵N provides moderate predictive power, with both sensitivity and specificity lower than δ¹³C and wider CI, reflecting higher uncertainty in classification.

### δ¹⁸O (`normalised_d18o`)

-   **Mean threshold:** 0.282 (CI: 0.201–0.353)\
-   **Sensitivity (AU correctly identified):** 0.86 (CI: 0.857–0.857)\
-   **Specificity (NZ correctly identified):** 0.71 (CI: 0.647–0.765)

**Interpretation:** δ¹⁸O is highly sensitive (good at detecting AU samples) but less specific (more false positives for NZ), with relatively narrow CI.

## General Interpretation

-   **δ¹³C** shows the best overall specificity (strong at identifying NZ samples).\
-   **δ¹⁸O** shows the best sensitivity (strong at identifying AU samples).\
-   The **multi-isotope model** balances sensitivity and specificity, leveraging complementary information from all three isotopes.\
-   Confidence intervals highlight **uncertainty due to small sample size**, but the bootstrap approach provides robust estimates of model performance and optimal thresholds.

**Summary:** Among the single isotopes, **δ¹³C is the best predictor** for geographic origin, providing the highest discriminative ability between NZ and AU samples. The sensitivity and specificity values indicate that the models are able to correctly classify a high proportion of both regions, with δ¹³C correctly identifying ~78% of AU samples and ~91% of NZ samples at the mean optimal threshold.

### C and N data from live birds
```{r}
sfc_as_cols <- function(x, geometry, names = c("x", "y")) {
  if (missing(geometry)) {
    geometry <- sf::st_geometry(x)
  } else {
    geometry <- rlang::eval_tidy(enquo(geometry), x)
  }
  
  stopifnot(inherits(x, "sf"), inherits(geometry, "sfc_POINT"))
  
  coords <- sf::st_coordinates(geometry)
  coords_df <- tibble::as_tibble(coords)
  
  stopifnot(length(names) == ncol(coords_df))
  
  x <- x[, !names(x) %in% names]
  coords_df <- setNames(coords_df, names)
  
  dplyr::bind_cols(x, coords_df)
}

bdot_phenotypes <- 
  read_excel(path = "/Users/luketheduke2/ownCloud/kemp_projects/bdot/R_projects/bdot_stable_isotopes/data/LIMS_251105_MaxPlanck_Eberhart-Hertel.xlsx",
             sheet = "Sample Submission Form", skip = 3,
             # range = cell_cols("A:Z"),
             col_names = TRUE) %>%
  select(`Sample Identifier`, TagID, Sex, `Individual Leg Markings`, `Migratory Status`, 
         `Date Sampled`, `Longitude Sampled`, `Latitude Sampled`, `Migratory Status Comment`) %>% 
  mutate(tag_id = ifelse(
    TagID == "266470" & as.Date(`Date Sampled`) < as.Date("2024-12-09"),
    "266470_A",
    ifelse(
      TagID == "266470" & as.Date(`Date Sampled`) >= as.Date("2024-12-09"),
      "266470_B",
      TagID
    )
  )) %>% 
  select(-c(TagID, `Date Sampled`)) %>% 
  rename(status = `Migratory Status`,
         latitude_cap = `Latitude Sampled`,
         longitude_cap = `Longitude Sampled`,
         comment = `Migratory Status Comment`,
         sex = Sex,
         mark = `Individual Leg Markings`,
         ring = `Sample Identifier`) %>% 
  mutate(ring = stringr::str_extract(ring, "^[^_]+")) %>% distinct()

live_bdot_data <-
  # read_excel(here("data/RAW DATA_SI_Feathers_LEH_Nov25.xls"), 
  #            sheet = "RAW_sample data",
  #            col_types = "text", skip = 9, col_names = TRUE) %>%
  read_excel(here("data/Master_EA_C_N_SI_Batch 5 Breast and Primary Feathers_LEH_Repeats added_April25_v1_SB_v2.xls"),
             sheet = "Sample results_C&N",
             col_types = "text", skip = 10, col_names = TRUE) %>%
    rename_with(~ .x %>%
                tolower() %>%
                str_replace_all(" ", "_") %>%
                str_replace_all("__", "_") %>%
                str_replace_all("___", "_") %>%
                str_replace_all("%", "_perc_") %>%
                str_replace_all(":", "") %>%
                str_replace_all("#", "")) %>% 
  dplyr::select(corrected_identifer_1, `amount_(mg)`, normalised_d15n, 
                normalised_d13c, cn_mass_ratio) %>% 
  rename(d15N = normalised_d15n,
         d13C = normalised_d13c,
         sample = `corrected_identifer_1`,
         amount = `amount_(mg)`) %>% 
  mutate(
    feather_type = str_extract(sample, regex("primary|breast|pimary", ignore_case = TRUE)),
    ring  = str_extract(sample, "^[^_]+")
  ) %>%
    # filter(ring != "CP19311") %>% 
  mutate(feather_type = ifelse(feather_type == "pimary", "primary", feather_type)) %>% 
  dplyr::select(ring, sample, feather_type, amount, d15N, d13C) %>% 
  mutate(amount = as.numeric(amount), 
         d15N = as.numeric(d15N), 
         d13C = as.numeric(d13C),
         d15N_scaled = as.numeric(scale(d15N)),
         d13C_scaled = as.numeric(scale(d13C))) %>%
  arrange(ring, feather_type) %>%
  # select(ring, feather_type, d15N, d13C) %>%
  # pivot_wider(
  #   names_from  = feather_type,
  #   values_from = c(d15N, d13C),
  #   names_glue  = "{.value}_{feather_type}"
  # ) %>% 
  left_join(., bdot_phenotypes, by = "ring") %>%
  mutate(status_bin = ifelse(str_detect(status, "migrant"), "migrant", status) %>% as.factor()) %>%
  filter(!is.na(ring)) %>% 
  filter(status != "unknown")
  #   na.omit() %>% 
  #   mutate(d15N_diff = abs(d15N_primary - d15N_breast),
  #          d13C_diff = abs(d13C_primary - d13C_breast)) %>% 
  # pivot_longer(
  #   cols = c(d15N_diff, d13C_diff),
  #   names_to = "isotope_diff",
  #   values_to = "value"
  # ) %>%
  # group_by(isotope_diff) %>%
  # mutate(value_scaled = as.numeric(scale(value))) %>%
  # ungroup() %>% 
  # left_join(., bdot_phenotypes, by = "ring") %>% 
  # mutate(status_bin = ifelse(str_detect(status, "migrant"), "migrant", status) %>% as.factor()) #%>% 
  # filter(status != "unknown")
bdot_argos_proc_2024 <- readRDS(file = "/Users/luketheduke2/ownCloud/kemp_projects/bdot/R_projects/bdot_db/output/bdot_argos_proc_2024_lonE.rds")

# Make midpoint the geometry and set it active
st_geometry(bdot_argos_proc_2024$ctdf) <- "midpoint"
library(geosphere)
bdot_winter_sites_2024_max_dist <- 
  bdot_argos_proc_2024$ctdf %>% 
  st_transform(4326) %>% 
  sfc_as_cols(names = c("longitude", "latitude")) %>% 
  rename(tag_id = id) %>% 
  # left_join(., bdot_phenotypes, by = "tag_id") %>% 
  st_drop_geometry() %>% 
  select(-chull) %>%
  mutate(status = ifelse(tag_id %in% c("266388", "266451"), "resident", status),
         longitude_cap = ifelse(tag_id == "266388", 172.6316, 
                                ifelse(tag_id == "266388", 172.6795, longitude_cap)),
         latitude_cap = ifelse(tag_id == "266388", -43.82988, 
                                ifelse(tag_id == "266388", -43.82687, latitude_cap))) %>% 
  mutate(distance_km = distHaversine(
    cbind(longitude, latitude),
    cbind(longitude_cap, latitude_cap)
  ) / 1000) %>% 
  group_by(tag_id) %>% 
  arrange(desc(distance_km)) %>% 
  slice(1) %>% 
  ungroup

bdot_winter_sites_2024_max_dist %>% 
  filter(status != "unknown") %>% 
  arrange(status)

live_bdot_data %>% 
  left_join(., 
            bdot_winter_sites_2024_max_dist %>% 
              select(ring, longitude, latitude, distance_km), 
            by = "ring") %>% 
  rename(longitude_winter = longitude, 
         latitude_winter = latitude,
         longitude_breed = longitude_cap,
         latitude_breed = latitude_cap) %>% 
  arrange(ring, feather_type) %>% 
  select(ring, feather_type, d15N, d13C, d15N_scaled, d13C_scaled, sex, status, 
         status_bin, longitude_breed, latitude_breed)#, longitude_winter, latitude_winter)

### Exploratory plot ----
ggplot() + 
  geom_line(data = live_bdot_data,
             aes(x = d13C_scaled, y = d15N_scaled, 
                 color = status, group = ring), 
             size = 0.5, alpha = 0.5) + 
  geom_point(data = live_bdot_data,
             aes(x = d13C_scaled, y = d15N_scaled, 
                 color = status, shape = feather_type), 
             size = 3) + 
  labs(x = expression(delta^{13}*C[scaled]), 
       y = expression(delta^{15}*N[scaled]), 
       shape = "Feather",
       color = "Status") + 
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "right", legend.title = element_blank(),
        axis.title = element_text(size = 15)) +
  facet_wrap(status ~ ., nrow = 2, ncol = 2)

df_dist <- 
  live_bdot_data %>%
  group_by(ring, status) %>%
  filter(n() == 2) %>%              # keep only groups with 2 points
  summarise(
    distance = sqrt((d15N_scaled[2] - d15N_scaled[1])^2 + (d13C_scaled[2] - d13C_scaled[1])^2)
  ) %>%
  ungroup() %>% 
  left_join(., bdot_winter_sites_2024_max_dist %>% select(ring, distance_km), by = "ring")

ggplot(data = df_dist) +
  geom_point(aes(x = distance, y = distance_km)) +
  geom_smooth(aes(x = distance, y = distance_km), method = "lm") +
  luke_theme +
  xlab("Eucledian distance in scaled N & C space") +
  ylab("Geographic distance between\nbreeding and wintering site (km)")

hist(df_dist$distance)

m1 <- aov(distance ~ status, data = df_dist)
summary(m1)
  
# Primaries
# 17 AU migrants, 22 residents, X unknown, 7 NI migrants, 10 SI migrants
live_bdot_data %>% 
  group_by(status) %>% summarise(n_birds = n_distinct(ring))

# 56 birds
live_bdot_data %>% 
  summarise(n_birds = n_distinct(ring))

# 55 birds with both feathers analyzed
live_bdot_data %>% 
  group_by(ring) %>% summarise(n_feathers = n_distinct(feather_type)) %>% 
  pull(n_feathers) %>% table()

live_bdot_data %>% 
  filter(!is.na(d15N) & !is.na(d13C)) %>% 
  group_by(feather_type) %>% summarise(n_birds = n_distinct(ring))

# Breast
# 11 AU migrants, 13 residents, 13 unknown, 5 NI migrants, 3 SI migrants
live_bdot_data %>% 
  filter(!is.na(d15N_breast) & !is.na(d13C_breast)) %>% 
  group_by(status) %>% summarise(n_birds = n_distinct(ring))

AUmig_res_primary <- 
  live_bdot_data %>% 
  filter(!is.na(d15N_primary) & !is.na(d13C_primary))
  
C_live_mod1 <- 
  glm(status_bin ~ value_scaled, 
       data = live_bdot_data %>% 
        filter(isotope_diff == "d13C_diff"), 
      family = binomial)

N_live_mod1 <- 
  glm(status_bin ~ value_scaled, 
       data = live_bdot_data %>% 
        filter(isotope_diff == "d15N_diff"), 
      family = binomial)

summary(C_live_mod1)
summary(N_live_mod1)

# 1️⃣ Create new data for prediction
newdataN <- 
  live_bdot_data %>% 
  filter(isotope_diff == "d15N_diff") %>% 
  reframe(
    value_scaled = seq(min(value_scaled), max(value_scaled), length.out = 50)
  ) %>%
  ungroup()

newdataC <- 
  live_bdot_data %>% 
  filter(isotope_diff == "d13C_diff") %>% 
  reframe(
    value_scaled = seq(min(value_scaled), max(value_scaled), length.out = 50)
  ) %>%
  ungroup()

# 2️⃣ Predict probabilities AND standard errors for CI
predC <- predict(C_live_mod1, newdataC, type = "link", se.fit = TRUE)
predN <- predict(N_live_mod1, newdataN, type = "link", se.fit = TRUE)

# Calculate predicted probability and 95% CI
newdataC <- newdataC %>%
  mutate(
    fit_link = predC$fit,
    se_link = predC$se.fit,
    # logistic transformation
    pred_prob = plogis(fit_link),
    lower = plogis(fit_link - 1.96 * se_link),
    upper = plogis(fit_link + 1.96 * se_link)
  )

newdataN <- newdataN %>%
  mutate(
    fit_link = predN$fit,
    se_link = predN$se.fit,
    # logistic transformation
    pred_prob = plogis(fit_link),
    lower = plogis(fit_link - 1.96 * se_link),
    upper = plogis(fit_link + 1.96 * se_link)
  )

# 3️⃣ Plot raw data + predicted probability + confidence intervals
ggplot() +
  # raw data
  geom_jitter(data = live_bdot_data %>% filter(isotope_diff == "d15N_diff"), 
              aes(x = value_scaled, y = status_bin, color = isotope_diff), 
              width = 0.1, height = 0.02, alpha = 0.6) +
  # predicted probability curves
  geom_line(data = newdataN, aes(x = value_scaled, y = pred_prob), size = 1.2) +
  # confidence interval ribbons
  geom_ribbon(data = newdataN, aes(x = value_scaled, ymin = lower, ymax = upper), 
              alpha = 0.2, color = NA) +
  theme_minimal() +
  labs(
    x = "Scaled difference in d13N",
    y = "Probability of having migrated"
  )
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# --- Prepare single-isotope predicted data ---
predict_iso <- function(mod, data, iso_name){
  pred <- predict(mod, se.fit = TRUE)
  data %>%
    ungroup() %>%
    mutate(
      linear_pred = pred$fit,
      fit = plogis(pred$fit),
      lower = plogis(pred$fit - 1.96*pred$se.fit),
      upper = plogis(pred$fit + 1.96*pred$se.fit),
      isotope = iso_name
    )
}

ggplot(data = live_bdot_data) +
  geom_boxplot(aes(x = status_bin, y = d15N, color = feather_type))

ggplot(data = live_bdot_data) +
  geom_boxplot(aes(x = status_bin, y = d13C, color = feather_type))

# δ13C
mod_d13C <- lm(d13C_scaled ~ status_bin, data = live_bdot_data)
pred_d13C <- predict_iso(mod_d13C, live_bdot_data, "δ13C")

# δ15N
mod_d15N <- lm(d15N_scaled ~ status_bin, data = live_bdot_data)
pred_d15N <- predict_iso(mod_d15N, live_bdot_data, "δ15N")

pred_all <- bind_rows(
  pred_d13C %>% dplyr::select(d13C_scaled, fit, lower, upper, isotope) %>% rename(mean_value = d13C_scaled),
  pred_d15N %>% dplyr::select(d15N_scaled, fit, lower, upper, isotope) %>% rename(mean_value = d15N_scaled)
)

# # δ15N + δ13C
# mod_multi <- glm(geo_region_bin ~ d13C + d15N,
#                  data = iso_wide, family = binomial)
# 
# pred_multi <- predict(mod_multi, type = "link", se.fit = TRUE)
# 
# iso_wide_pred <- iso_wide %>%
#   mutate(
#     linear_pred = pred_multi$fit,
#     fit = plogis(pred_multi$fit),
#     lower = plogis(pred_multi$fit - 1.96*pred_multi$se.fit),
#     upper = plogis(pred_multi$fit + 1.96*pred_multi$se.fit),
#     isotope = "multi-isotope"
#   )

# --- Combine all predicted data ---
# pred_all <- bind_rows(
#   pred_d13c %>% dplyr::select(mean_value, fit, lower, upper, isotope),
#   pred_d15n %>% dplyr::select(mean_value, fit, lower, upper, isotope),
#   pred_d18o %>% dplyr::select(mean_value, fit, lower, upper, isotope),
#   pred_d2h %>% dplyr::select(mean_value, fit, lower, upper, isotope)#,
#   # iso_wide_pred %>% dplyr::select(linear_pred, fit, lower, upper, isotope) %>%
#     # rename(mean_value = linear_pred)
# )

# Combine raw data points for plotting
raw_all <- bind_rows(
  live_bdot_data %>% dplyr::select(d13C_scaled, status_bin) %>% mutate(isotope = "δ13C") %>% rename(mean_value = d13C_scaled),
  live_bdot_data %>% dplyr::select(d15N_scaled, status_bin) %>% mutate(isotope = "δ15N") %>% rename(mean_value = d15N_scaled)#,
  # d18o_data %>% dplyr::select(mean_value, geo_region_bin) %>% mutate(isotope = "δ18O"),
  # d2h_data %>% dplyr::select(mean_value, geo_region_bin) %>% mutate(isotope = "δ2H")#,
  # iso_wide_pred %>% dplyr::select(linear_pred, geo_region_bin) %>% mutate(isotope = "multi-isotope") %>%
    # rename(mean_value = linear_pred)
)

# Thresholds for plotting
thresholds <- data.frame(
  isotope = c("δ13C","δ15N","δ18O","δ2H"), #,"multi-isotope"),
  threshold = c(0.475, 0.366, 0.282, 0.59)
)

# --- Plot ---
ggplot() +
  geom_point(data = raw_all, aes(x = mean_value, y = as.numeric(status_bin)-1),
             position = position_jitter(height = 0.05), alpha = 0.6) +
  geom_line(data = pred_all, aes(x = mean_value, y = fit), color = "steelblue", size = 1) +
  geom_ribbon(data = pred_all, aes(x = mean_value, ymin = lower, ymax = upper),
              fill = "steelblue", alpha = 0.2) +
  # geom_hline(data = thresholds, aes(yintercept = threshold), linetype = "dashed", color = "red") +
  facet_wrap(~isotope, scales = "free_x") +
  labs(x = "Stable isotope value",
       y = "Probability wintering in Australia",
       title = "Predicted effects of single-isotope models on wintering region") +
  theme_minimal()
```

```{r}
# Required packages
packages <- c("tidyverse","tidymodels","caret","lme4","brms","MASS","randomForest","pROC","patchwork","rsample","vip")
invisible(lapply(packages, function(p) if(!requireNamespace(p, quietly=TRUE)) install.packages(p)))
library(tidyverse); library(tidymodels); library(caret); library(lme4); library(brms)
library(MASS); library(randomForest); library(pROC); library(patchwork); library(rsample); library(vip)
library(DBI); library(dbo); library(elevatr); library(gghalves)
library(dbo)
library(elevatr)
con = dbcon(db = "FIELD_2024_BADOatNZ")

df <- 
  live_bdot_data %>% 
  left_join(., 
            bdot_winter_sites_2024_max_dist %>% 
              select(ring, longitude, latitude, distance_km), 
            by = "ring") %>% 
  rename(longitude_winter = longitude, 
         latitude_winter = latitude,
         longitude_breed = longitude_cap,
         latitude_breed = latitude_cap) %>% 
  arrange(ring, feather_type) %>% 
  select(ring, feather_type, d15N, d13C, d15N_scaled, d13C_scaled, sex, status, 
         status_bin, longitude_breed, latitude_breed) %>% 
  left_join(., dbq(con, "SELECT * FROM CAPTURES") %>% select(ring, site) %>% distinct()) %>% 
  group_by(ring) %>% 
  mutate(n_tissues = n_distinct(feather_type)) %>% 
  ungroup() %>% 
  mutate(x_line = ifelse(feather_type == "primary", 1 + 0.15,
                         ifelse(feather_type == "breast", 
                                2 - 0.15, NA)),
         x_point = ifelse(feather_type == "primary", 1, 
                          ifelse(feather_type == "breast", 
                                 2, NA)),
         x_box = ifelse(feather_type == "primary", 1 - 0.15, 
                          ifelse(feather_type == "breast", 
                                 2 + 0.15, NA))) %>% 
  mutate(status = factor(status, levels = c("resident", "SI migrant", "NI migrant", "AU migrant"))) %>% 
  mutate(site_name = case_when(
      site == "KT" ~ "Kaitorete Spit",
      site == "TS" ~ "Tasman River",
      site == "KP" ~ "Kena Peninsula",
      site == "MS" ~ "Motueka Spit",
      site == "KK" ~ "Kaikoura",
      site == "CR" ~ "Cass River",
      site == "TA" ~ "Upukeroa River",
      site == "TP" ~ "Tiwai Point",
      site == "CH" ~ "Chathams",
      site == "HC" ~ "Napier",
      site == "OM" ~ "Old Man Range",
      site == "PR" ~ "Pisa Range",
      site == "PB" ~ "Porters Beach, Abel Tasman National Park",
      site == "TO" ~ "Rangipo Desert",
      site == "MC" ~ "Macaulay River",
      site == "MB" ~ "Maryburn",
      site == "OD" ~ "Ohau Downs",
      site == "TR" ~ "Tekapo River",
      site == "MA" ~ "Miranda Rescue",
      site == "PW" ~ "Port Waikato",
      TRUE ~ site
    ))

status_facets <- c('SI migrant' = "South Island migrant", 
                   'NI migrant' = "North Island migrant", 
                   'AU migrant' = "Australia migrant", 
                   'resident' = "Resident")

# Step 1: Convert to sf object
points_sf <- st_as_sf(df, coords = c("longitude_breed", "latitude_breed"), crs = 4326)

# Step 2: Get elevation (units: meters)
elevations <- get_elev_point(locations = points_sf, units = "meters", src = "aws")

# Step 3: Add elevation to original data
df <- 
  bind_cols(df, elevation = elevations$elevation) %>% 
  mutate(elevation = ifelse(elevation < 0, 0, elevation))

df_wide <- df %>% 
  select(ring, feather_type, d15N, elevation, status) %>% 
  group_by(ring, elevation, status, feather_type) %>% 
  summarise(d15N = mean(d15N, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(
    names_from = feather_type,
    values_from = d15N
  ) %>% 
  mutate(
    d15N_diff = primary - breast
  ) %>% 
  drop_na(d15N_diff)# %>% 
  # filter(status %in% c("AU migrant", "NI migrant"))

mod <- lm(d15N_diff ~ elevation, data = df_wide)
summary(mod)

newdat <- expand.grid(
  elevation = seq(min(df_wide$elevation), max(df_wide$elevation), length.out = 100),
  status = unique(df_wide$status)
)

pred <- predict(mod, newdata = newdat, interval = "confidence")
newdat$fit <- pred[, "fit"]
newdat$lwr <- pred[, "lwr"]
newdat$upr <- pred[, "upr"]

ggplot() +
  geom_point(data = df_wide,
             aes(x = elevation, y = d15N_diff),
             alpha = 0.7) +
  geom_ribbon(data = newdat,
              aes(x = elevation, ymin = lwr, ymax = upr),
              alpha = 0.25, colour = NA) +
  geom_line(data = newdat,
            aes(x = elevation, y = fit),
            size = 1.2) +
  labs(
    x = "Elevation",
    y = "Difference in d15N (primary – breast)",
    color = "Status"
  ) +
  theme_minimal(base_size = 14)# +
  # facet_grid(.~status)

# ---------------------------
# 0. Basic data assumptions
# ---------------------------
# Input: one row per feather (like your sample). Columns required:
# ring (unique bird id), feather_type (e.g., "primary" or "breast"),
# d15N, d13C, sex, status (resident/migrant), longitude_breed, latitude_breed
# If you have scaled values already, you can use them (d15N_scaled etc.) or compute within pipeline.

# Example: df <- read_csv("your_data.csv")
# I'll assume df exists.

# ---------------------------
# 1. Reshape to one-row-per-bird
# ---------------------------
# We need paired samples per bird (primary = breeding feather, breast = winter feather).
# If some birds miss one feather, we can decide to keep them later (but paired data is best).

df_wide <- df %>%
  group_by(site, ring, sex, status, longitude_breed, latitude_breed, elevation, feather_type) %>%
  summarise(
    d15N = mean(d15N, na.rm = TRUE),
    d13C = mean(d13C, na.rm = TRUE),
    d15N_scaled = mean(d15N_scaled, na.rm = TRUE),
    d13C_scaled = mean(d13C_scaled, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = feather_type,
              values_from = c(d15N, d13C, d15N_scaled, d13C_scaled),
              names_sep = "_") %>%
  # Derive new useful cols: deltas (breast - primary), and absolute delta
  mutate(
    delta_d15N = d15N_breast - d15N_primary,
    delta_d13C = d13C_breast - d13C_primary,
    abs_delta_d13C = abs(delta_d13C),
    abs_delta_d15N = abs(delta_d15N),
    # keep factor status with resident as reference
    status = factor(status, levels = c("resident", "SI migrant", "NI migrant", "AU migrant"))
  )
# Check number of complete pairs
n_pairs <- df_wide %>% filter(!is.na(d15N_primary) & !is.na(d15N_breast)) %>% nrow()
message("Paired birds: ", n_pairs)

# ---------------------------
# 2. Exploratory plots
# ---------------------------
# Scatter primary vs breast for each isotope; and delta histograms
p1 <- ggplot(df_wide, aes(x = d13C_primary, y = d13C_breast, colour = status)) +
  geom_abline(lty=2) + 
  geom_point(size=3, alpha=0.8) + 
  labs(x="d13C primary (breeding)", y="d13C breast (winter)")

p2 <- ggplot(df_wide, aes(x = d15N_primary, y = d15N_breast, colour = status)) +
  geom_abline(lty=2) + 
  geom_point(size=3, alpha=0.8) + 
  labs(x="d15N primary", y="d15N breast")

p3 <- ggplot(df_wide, aes(x = delta_d13C, fill = status)) + 
  geom_density(alpha=0.4) + 
  labs(x="delta d13C (breast - primary)")

p4 <- ggplot(df_wide, aes(x = delta_d15N, fill = status)) + 
  geom_density(alpha=0.4) + 
  labs(x="delta d15N (breast - primary)")

(p1 + p2) / (p3 + p4)

### box plots of status and feather type ----
# p5 <- 
  ggplot(df %>% filter(n_tissues == 2) %>% filter(status %in% c("AU migrant", "resident")), 
         aes(x = status, y = d15N, fill = feather_type)) + 
  geom_boxplot(alpha = 0.4) +
  theme(#legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# p6 <- 
  ggplot(df %>% filter(n_tissues == 2) %>% filter(status %in% c("AU migrant", "resident")), 
         aes(x = status, y = d13C, fill = feather_type)) + 
  geom_boxplot(alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p7 <-
  ggplot() +
  geom_line(data = df %>% filter(n_tissues == 2), 
            aes(x = x_line, y = d13C, group = ring, color = site_name)) +
  geom_jitter(data = df %>% filter(n_tissues == 2),
              size = 4, aes(x = x_point, y = d13C, fill = site_name),
              shape = 21, alpha = 1, width = 0.075) +
  geom_half_boxplot(data = df %>% filter(n_tissues == 2) %>% filter(feather_type == "primary"), 
                    aes(x = x_box, y = d13C), side = "l") +
  geom_half_boxplot(data = df %>% filter(n_tissues == 2) %>% filter(feather_type == "breast"), 
                    aes(x = x_box, y = d13C), side = "r") +
  facet_grid(~ status, labeller = as_labeller(status_facets)) +
  theme_light() +
  theme(legend.position = "top",
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        strip.text = element_text(size = 11)) +
  scale_x_continuous(limits = c(0.3, 2.7), breaks = c(1, 2), labels = c("Primary", "Breast")) +
  xlab("Feather tissue") +
  labs(color = NULL, fill = NULL, size = NULL)

p8 <-
  ggplot() +
  geom_line(data = df %>% filter(n_tissues == 2), 
            aes(x = x_line, y = d15N, group = ring, color = site_name)) +
  geom_jitter(data = df %>% filter(n_tissues == 2),
              size = 4, aes(x = x_point, y = d15N, fill = site_name),
              shape = 21, alpha = 1, width = 0.075) +
  geom_half_boxplot(data = df %>% filter(n_tissues == 2) %>% filter(feather_type == "primary"), 
                    aes(x = x_box, y = d15N), side = "l") +
  geom_half_boxplot(data = df %>% filter(n_tissues == 2) %>% filter(feather_type == "breast"), 
                    aes(x = x_box, y = d15N), side = "r") +
  facet_grid(~ status, labeller = as_labeller(status_facets)) +
  theme_light() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_blank()) +
  scale_x_continuous(limits = c(0.3, 2.7), breaks = c(1, 2), labels = c("Primary", "Breast")) +
  xlab("Feather tissue")

(p7 / p8)
```

```{r}
library(MASS)
# Remove rows with any missing isotopes
iso_lda <- df %>%
  drop_na(d15N_scaled, d13C_scaled) %>% 
  mutate(geo_region_bin = ifelse(status == "AU migrant", 1, 0))

# Make geo_region_bin a factor (required by lda)
iso_lda <- iso_lda %>%
  mutate(region = factor(geo_region_bin, levels = c(0, 1), labels = c("NZ", "AU")))

# Run LDA
lda_model <- lda(region ~ d13C_scaled + d15N_scaled, data = iso_lda)

# View results
lda_model

# Get predictions
pred <- predict(lda_model)

# Add to dataframe
iso_lda <- iso_lda %>%
  mutate(LD1 = pred$x[, 1],
         predicted = pred$class)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = iso_lda$predicted)

# Classification accuracy
mean(iso_lda$region == iso_lda$predicted)

ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "LDA separation between AU and NZ individuals",
       x = "Linear Discriminant 1", fill = "Region")

ggplot(iso_lda, aes(x = LD1, y = 0, color = region)) +
  geom_jitter(width = 0, height = 0.05, size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(x = "LD1", y = NULL, title = "LDA projection")

lda_cv <- lda(region ~ d13C_scaled + d15N_scaled, data = iso_lda, CV = TRUE)

# Accuracy
mean(lda_cv$class == iso_lda$region)

# Confusion matrix
table(Observed = iso_lda$region, Predicted = lda_cv$class)

ggplot(iso_lda, aes(x = region, y = LD1, color = region)) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", size = 4, shape = 23, fill = "white") +
  scale_color_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  labs(title = "LDA Separation by Region",
       y = "LD1 score",
       x = "Region")

# Assuming lda_model is already fitted
lda_coeff <- lda_model$scaling[, 1]  # LD1 coefficients
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_coeff
) %>%
  arrange(desc(abs(coefficient)))  # order by absolute contribution

ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = "none") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- LDA model (already fitted) ---
lda_model <- lda(region ~ d13C + d15N + d2H + d18O, data = iso_lda)

# --- LD1 scores ---
lda_pred <- predict(lda_model)
lda_df <- data.frame(CNOH_wide, LD1 = lda_pred$x[, 1])

# --- LD1 density plot ---
p1 <- ggplot(iso_lda, aes(x = LD1, fill = region)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("AU" = "#E69F00", "NZ" = "#56B4E9")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "LDA separation by region",
    x = "LD1",
    y = "Density",
    fill = "Region"
  )

# --- Extract LD1 coefficients (loadings) ---
lda_coeff_df <- data.frame(
  isotope = rownames(lda_model$scaling),
  coefficient = lda_model$scaling[, 1]
) %>%
  arrange(desc(abs(coefficient)))

# --- Isotope contribution bar plot ---
p2 <- ggplot(lda_coeff_df, aes(x = reorder(isotope, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"), guide = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Contribution of isotopes to LD1",
    x = "Isotope",
    y = "LD1 coefficient"
  )

# --- Combine plots with patchwork ---
combined_plot <- p1 / p2 + plot_layout(heights = c(2, 1))
combined_plot
```


```{r}
# Also pairwise PCA (optional)
df_pca <- df_wide %>%
  select(d13C_primary, d13C_breast, d15N_primary, d15N_breast) %>%
  na.omit()
pca_res <- prcomp(df_pca %>% scale(), center = TRUE)
autoplot_pca <- function() {
  p <- tibble(pca1 = pca_res$x[,1], pca2 = pca_res$x[,2], status = df_wide$status[!is.na(df_wide$d13C_primary & df_wide$d13C_breast)]) %>%
    ggplot(aes(pca1,pca2,colour=status)) + geom_point(size=3)
  p
}
print(autoplot_pca())

# ---------------------------
# 3. Feature engineering choices
# ---------------------------
# Candidate predictors (try several sets in models):
#  A) primary feather isotopes (d13C_primary, d15N_primary)
#  B) breast feather isotopes (d13C_breast, d15N_breast)
#  C) deltas (delta_d13C, delta_d15N)
#  D) combinations (primary + delta)
# Also consider: sex, breeding longitude/latitude, altitude or site (if you have a site id),
# and interactions (e.g. sex * delta_d13C)
#
# Standardize predictors for models that need it:
num_vars <- c("d13C_primary","d15N_primary","d13C_breast","d15N_breast","delta_d13C","delta_d15N","abs_delta_d13C","abs_delta_d15N")
df_wide <- df_wide %>%
  mutate(across(all_of(num_vars), ~ scale(.)[,1], .names = "z_{col}"))

# ---------------------------
# 4. Cross-validation strategy
# ---------------------------
# --- diagnostics: check site counts first ---
library(dplyr)
n_sites <- df_wide %>% filter(!is.na(delta_d13C)) %>% pull(site) %>% unique() %>% length()
site_counts <- df_wide %>% filter(!is.na(delta_d13C)) %>% count(site, sort = TRUE)

message("Unique sites with paired data: ", n_sites)
print(head(site_counts, 20))

# If you have fewer than 2 sites, grouped CV doesn't make sense:
if(n_sites < 2) stop("Need at least 2 unique sites for grouped CV. Found: ", n_sites)

# choose v sensibly: at most number of sites, default 5 but reduce if needed
v_desired <- 5L
v <- min(v_desired, n_sites)
if(v < v_desired) message("Reduced v to number of sites: v = ", v)

# remove rows missing the predictors you need (e.g. delta_d13C) before cv
df_for_cv <- df_wide %>% filter(!is.na(delta_d13C) & !is.na(d13C_primary) & !is.na(d13C_breast))

# --- Correct usage: pass the column name (unquoted) or as string ---
# Option A: unquoted column name (tidy-eval)
set.seed(123)
group_folds <- rsample::group_vfold_cv(df_for_cv, group = site, v = v)

# Option B: pass column name as a string
# group_folds <- rsample::group_vfold_cv(df_for_cv, group = "site", v = v)

# --- Quick check of folds produced ---
# each element of group_folds is a vfold object; inspect the first split
group_folds$splits[[1]] %>% print()
# get the grouping sites left-out in the first split
first_assessment <- assessment(group_folds$splits[[1]])
leftout_sites <- setdiff(df_for_cv$site, first_assessment$site) # not quite right; better to compute directly:
# better: we can get the analysis set and assessment set and see which sites are in the assessment:
analysis_sites <- unique(analysis(group_folds$splits[[1]])$site)
assessment_sites <- unique(assessment(group_folds$splits[[1]])$site)
message("First split: analysis sites = ", length(analysis_sites), "; assessment sites = ", length(assessment_sites))
message("Example assessment sites (first split): ", paste(head(assessment_sites, 10), collapse = ", "))

# Optional: if some sites have only 1 bird and you want to avoid leaving them alone in assessment,
# consider grouping by larger unit or grouping rare singletons together.

# Print a summary of number of rows in analysis & assessment for each split
fold_summary <- map_dfr(group_folds$splits, ~ tibble(
  id = attr(.x, "id"),
  n_analysis = nrow(analysis(.x)),
  n_assessment = nrow(assessment(.x)),
  n_sites_analysis = n_distinct(analysis(.x)$site),
  n_sites_assessment = n_distinct(assessment(.x)$site)
))
print(fold_summary)

# Because birds may cluster by site, use grouped CV where each fold leaves out entire sites.
# If you have a column for site or geo_region, use that. I'll use longitude_breed rounded as a proxy.
df_wide <- df_wide %>% mutate(site = as.character(site))

set.seed(123)
# grouped 5-fold CV; replace site_group with 'site' if you have it
group_folds <- group_vfold_cv(df_wide %>% filter(!is.na(delta_d13C)), group = df_wide$site, v = 5)

# ---------------------------
# 5. Modeling with tidymodels (logistic, LDA, random forest)
# ---------------------------
# We'll build a function to evaluate models across folds and compute metrics.

# Helper: formula for model (example: using deltas & primary)
model_formula <- status ~ d13C_primary + d15N_primary + delta_d13C + delta_d15N


df_wide$status_bin <- ifelse(df_wide$status == "resident", "resident", "migrant")
df_wide$status_bin <- factor(df_wide$status_bin)

# Prepare recipe
rec <- recipe(status_bin ~ d13C_primary + d15N_primary + delta_d13C + delta_d15N + ring, 
              data = df_wide) %>%
  update_role(ring, new_role = "ID") %>%
  # step_string2factor(sex) %>%
  step_normalize(all_numeric_predictors())

# Logistic Regression (multinomial if >2 classes)
log_spec <- multinom_reg(mode = "classification") %>%
  set_engine("nnet")   # robust multinomial implementation

# Linear Discriminant Analysis
lda_spec <- discrim_linear() %>%
  set_engine("MASS")

# Random Forest
rf_spec <- rand_forest(
  trees = 1000,
  mtry = 3,
  min_n = 5
) %>% 
  set_mode("classification") %>%
  set_engine("ranger", importance = "impurity")

# logistic workflow
wf_log <- workflow() %>%
  add_recipe(rec) %>%
  add_model(log_spec)
library(discrim)
# install.packages("discrim")
# LDA workflow
wf_lda <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lda_spec)

# random forest workflow
wf_rf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_spec)

# Fit with resampling
set.seed(123)
folds <- vfold_cv(df_wide, v = 10, strata = status_bin)

# install.packages("finetune")
library(finetune)

# Function to evaluate workflows
eval_model <- function(wflow) {
  fit_res <- fit_resamples(
    wflow,
    resamples = folds,
    metrics = metric_set(accuracy, roc_auc),
    control = control_resamples(save_pred = TRUE)
  )
  return(fit_res)
}
# install.packages("ranger")
library(ranger)
res_log <- eval_model(wf_log)
res_lda <- eval_model(wf_lda)
res_rf  <- eval_model(wf_rf)


log_res <- fit_resamples(wf_log, resamples = group_folds, metrics = metric_set(roc_auc, accuracy, sens, spec), control = control_resamples(save_pred = TRUE))
lda_res <- fit_resamples(wf_lda, resamples = group_folds, metrics = metric_set(roc_auc, accuracy, sens, spec), control = control_resamples(save_pred = TRUE))
rf_res  <- fit_resamples(wf_rf,  resamples = group_folds, metrics = metric_set(roc_auc, accuracy, sens, spec), control = control_resamples(save_pred = TRUE))

collect_metrics(log_res)
collect_metrics(lda_res)
collect_metrics(rf_res)

# Inspect pooled ROC AUC using saved predictions (example for logistic):
log_preds <- collect_predictions(log_res)
roc_auc(log_preds, truth = status, .pred_migrant) # .pred_migrant will exist if glm returns prob for class 'migrant'

# ---------------------------
# 6. Evaluate best model & calibration
# ---------------------------
# Pick winner (e.g. rf_res). Refit on full dataset (or fit a brms GLMM if you want random effects)
best_wf <- wf_rf
final_fit <- last_fit(best_wf %>% update_model(rand_forest(trees = 1000) %>% set_engine("randomForest")), split = initial_split(df_wide %>% filter(!is.na(delta_d13C)), strata = status))
# or simply fit to whole dataset
final_model <- fit(best_wf, data = df_wide)

# Variable importance (for RF)
rf_fit_obj <- extract_fit_parsnip(final_model)$fit
varImp <- importance(rf_fit_obj) %>% as.data.frame() %>% rownames_to_column("variable") %>% arrange(desc(MeanDecreaseGini))
print(varImp)

# Calibration plot for predicted probabilities
preds_all <- predict(final_model, df_wide, type = "prob") %>% bind_cols(df_wide %>% select(status))
preds_all <- preds_all %>% mutate(prob = .pred_migrant)
calib_df <- preds_all %>% mutate(bin = ntile(prob, 10)) %>% group_by(bin) %>% summarise(mean_pred = mean(prob), obs = mean(status == "migrant"), n = n())
ggplot(calib_df, aes(x = mean_pred, y = obs)) + geom_point() + geom_abline(slope = 1, intercept = 0, lty=2) + labs(x="Mean predicted prob (migrant)", y="Observed frequency")

# ---------------------------
# 7. Grouped (site-aware) performance summary
# ---------------------------
# It's often useful to show sensitivity/specificity by site (or region)
preds_by_fold <- collect_predictions(rf_res)
# compute confusion matrices per resample
conf_by_split <- preds_by_fold %>%
  group_by(id) %>%
  summarise(
    cm = list(conf_mat(data = ., truth = status, estimate = .pred_class))
  )
conf_by_split

# ---------------------------
# 8. Bayesian GLMM (optional, for uncertainty & site random effects)
# ---------------------------
# If you want to include random intercept for site and report posterior probability:
# Requires brms installed and a working backend; this can be slow but gives full posterior probs.

# Create site factor (replace with real site ID if available)
df_wide <- df_wide %>% mutate(site = site_group)
brms_formula <- bf(as.integer(status=="migrant") ~ d13C_primary + d15N_primary + delta_d13C + delta_d15N + sex + (1|site))
# Fit (short chains for example)
# fit_brms <- brm(brms_formula, data = df_wide, family = bernoulli(link="logit"), chains=4, cores=4, iter=2000)
# After fit, get posterior predicted probability for new observations with posterior_predict or posterior_linpred (type="response")

# ---------------------------
# 9. Predict function for a new bird
# ---------------------------
# Assume you will supply d13C_primary, d15N_primary, d13C_breast, d15N_breast (or deltas). This returns predicted probability of migrant.

predict_migration <- function(model_fit, newdata_df){
  # newdata_df should have columns: d13C_primary, d15N_primary, d13C_breast, d15N_breast, sex (if used)
  newdata_df <- newdata_df %>%
    mutate(delta_d13C = d13C_breast - d13C_primary,
           delta_d15N = d15N_breast - d15N_primary)
  probs <- predict(model_fit, newdata_df, type = "prob")
  # return dataframe with prob of migrant
  bind_cols(newdata_df, probs)
}

# Example usage:
# newbird <- tibble(d13C_primary = -21.0, d15N_primary = 12.3, d13C_breast = -17.0, d15N_breast = 14.1, sex = "F")
# predict_migration(final_model, newbird)

# ---------------------------
# 10. Thresholding & reporting
# ---------------------------
# Choose a classification threshold depending on your goal:
# - If missing a migrant is costly, choose threshold to maximize sensitivity.
# - If false positives are costly, maximize specificity.
# Use ROC curve to pick threshold:
rf_preds <- collect_predictions(rf_res)
roc_obj <- roc(response = rf_preds$truth, predictor = rf_preds$.pred_migrant)
coords(roc_obj, x="best", best.method = "youden") # returns threshold maximizing sensitivity+specificity

# ---------------------------
# 11. Quick realism assessment
# ---------------------------
# After running the models, assess:
# - AUC > 0.80 => good discrimination (screening feasible),
# - 0.7 < AUC < 0.8 => moderate (useful as a triage tool),
# - AUC < 0.7 => poor for single-bird classification.
# Also inspect confusion matrix, and site-wise performance. If performance differs strongly among sites, you may need site-specific models or include spatial covariates.

# ---------------------------
# 12. Sensitivity checks & suggestions
# ---------------------------
# - Try simpler rule-based classifier: e.g., if delta_d13C > X then migrant. Use ROC to pick X.
# - Test models restricted to alpine or coastal birds to see if patterns differ.
# - If you have sample-age or feather molt timing metadata, include it.
# - External validation: ideally test on an independent holdout of birds not used in training.
# - Bayesian GLMM recommended if you want full posterior uncertainty and to borrow strength across sites.
# - Beware possible trophic or dietary differences (d15N) that might confound movement signal; interpret with ecology in mind.

# End of pipeline
message("Pipeline script executed. Fit models and inspect metrics to judge feasibility.")

```

